### 2011-12-18 paper.tex
Placeholders placed

\documentclass[letterpaper, 12pt]{article}
%%\documentclass[letterpaper, 11pt, twocolumn]{article}
\usepackage{url}
\usepackage{ctable}
\usepackage{overcite}
\usepackage{amsmath,amssymb}      % for \AA etc
\usepackage[onehalfspacing]{setspace}
\usepackage{graphicx}
\usepackage{fullpage, times}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{longtable}

\DeclareGraphicsExtensions{.pdf,.png}
\onehalfspacing

\bibliographystyle{jcics}

\usepackage{anysize}
%\marginsize{1in}{2in}{1in}{1in}
%\marginparwidth 1.75in
%\newcommand{\mnote}[1]{\marginpar{\raggedright{\footnotesize\textbf{#1}}}}

\def\imagetop#1{\vtop{\null\hbox{#1}}}

%%%
%%% Don't change the size of the figures as they are defined for
%%% a single column environment, which is what we will be submitting
%%% as
%%%

\begin{document}
\title{Foo}
\author{Rajarshi Guha${}^{\dagger}$ and Egon Willighagen${}^{\ddagger}$\\
${}^{\dagger}$NIH Center for Translational Therapeutics\\9800 Medical
Center Drive  Rockville, MD 20850 \\
${}^{\ddagger}$ Somwhere in the Netherlands
}
\date{}

\maketitle
\begin{abstract}
Bar
\end{abstract}

\section{Introduction}

\clearpage
\newpage

\bibliography{paper}

\newpage

\end{document}

### 2012-02-29 paper.tex
Added some notes to build the article around. Added in a title

@@
 \begin{document}
-\title{Foo}
+\title{A Survey of Quantitative Descriptions of Molecular Structure}
 \author{Rajarshi Guha${}^{\dagger}$ and Egon Willighagen${}^{\ddagger}$\\
-${}^{\dagger}$NIH Center for Translational Therapeutics\\9800 Medical
+${}^{\dagger}$NIH Center for Advancing Translational Science\\9800 Medical
 Center Drive  Rockville, MD 20850 \\
@@
 
+\section{Running Notes}
+\begin{itemize}
+\item What are descriptors, why are they needed?
+\item Survey of descriptor types, focusing on nature of inputs,
+  complexity, interpretabilty 
+\item Talk about biological descriptors (see paper from Tropsha that
+  uses bioactivity data as descriptors)
+\item Discuss how fingerprints can also be used as descriptors
+\item Also mention descriptors for non-small molecule cases
+  \begin{itemize}
+  \item proteins
+  \item polymers
+  \end{itemize}
+\item Survey of descriptor implementations (OSS and non-OSS)
+  \begin{itemize}
+  \item MOE
+  \item DRAGON
+  \item CDK
+  \item RDKit
+  \item JChem
+  \item PipelinePilot
+  \item OEChem/GraphTK
+  \item JOELib
+  \item CODESSA
+  \item Schrodinger stuff (Meastro?)
+  \end{itemize}
+\item Go into more detail of OSS descriptor implementations
+  \begin{itemize}
+  \item CDK (describe architecture), mention wrappers around the
+    library (CDKDescUI and batch mode)
+  \item PaDEL
+  \end{itemize}
+\item Compare OSS descs to commercial ones - point to the Pfizer/CDD
+  paper that compared MOE and CDK
+\end{itemize}
 \clearpage

### 2012-03-10 paper.tex
Added my address

@@
 ${}^{\dagger}$NIH Center for Advancing Translational Science\\9800 Medical
-Center Drive  Rockville, MD 20850 \\
-${}^{\ddagger}$ Somwhere in the Netherlands
+Center Drive  Rockville, MD 20850 \\U.S.A.\\
+${}^{\ddagger}$ Department of Bioinformatics - BiGCaT\\Maastricht University\\P.O. Box 616\\NL-6200 MD Maastricht\\The Netherlands
 }

### 2012-03-15 paper.tex
Added some introduction

@@
 
+Computational methods play an important role in many chemical
+disciplines ranging from drug discovery to materials science. While
+there are a variety of techniques that differ in terms of
+computational complexity, time requirements and so on. However the
+common requirement underlying all these methods is a formal
+description of a the molecular structure. Of course, there are many
+ways to ``describe'' a molecule. A common approach is to describe the
+connectivity, taking into account the types of atoms and bonds. In
+other words, explicit representations of chemical structure such as
+SMILES, SD files and so on. While these descriptions are vital to
+modern chemical information systems, they do not necessarily allow
+computational techniques to be directly applied to them.
+
+Methods that aim to predict chemical and biological properties
+generally require a numerical description of chemical structures. Such
+numerical forms range from a set of 3D coordinates (which coupled with
+appropiate atom types, is sufficient for methods such QM approaches
+and docking) to more abstract numerical descriptions derived from 2D
+or 3D representations which can be useful is statistical approaches.
+
+It is now possible to evaluate thousands of numerical descriptors of
+chemical structure. As will be discussed later, many of these
+descriptors are closely related - one can be substituted for
+another. The choice of descriptors is a well known problem and given a
+large collection of them, approaches to identify a suitable subset
+have been discussed extensively in the literature [REFS].
+
+In addition to there being many possible descriptors defined in the
+literature, there are also multiple implementations of a give
+descriptor. These implementations are available in the form of
+libraries (which require one to write a program to use them) or
+complete applications (GUI or command line). As a result, not only
+must one choose one or more descriptors that are relevant to the
+problem at hand, but in addition, one must be concerned about how they
+are calculated and whether such a calculation can be reproduced across
+different implementations of those descriptors. It is easy to
+understand \emph{why} two implementations of the same descriptor can
+lead to different results - the primary reason being the chemistry
+model that underlies the framework or toolkit used to implement the
+descriptor. For example, a descriptor that calculates the number of
+aromatic atoms may be implemented using two toolkits with differing
+aromaticity models, and hence it is possible that the values generated
+by the two implementations will differ. Other sources of differences
+include parameters that may be involved in the descriptor calculation
+and reference data values (such as atomic radii, electronegativity
+values) that are employed during descriptor calculation. While most
+implementations will employ the same data sources for standard
+concepts (e.g., atomic weights), slight differences in these types of
+input data can lead to differences in the final descriptor value. As a
+result, in most cases, two implementations of a descriptor do not
+usually give the exact same value, though they are usually quite
+similar. Explicitly explaining the differences may or may not be
+possible (it is usually more difficult in the cases of commercial
+implementations for which source code is not available).
+
 \section{Running Notes}
@@
   \item polymers
+  \item inorganics (maybe ask Maciej for pointers)
   \end{itemize}

### 2012-03-16 paper.tex
Added a chunk of text and started added references

@@
 
+\section{A Categorization of Descriptors}
+\label{sec:categ-descr}
+
+We have noted that a molecular structure can be characterized using a
+numebr of numerical descriptors. In this section we describe a
+categorization of descriptors. Admittedly, the grouping is somewhat
+arbitrary but serves as a useful guide. In addition, the
+categorization also considers the nature of the chemical structure
+being considered - some descriptors are only useful when applied to
+small molecules, whereas other descriptors are defined specifically
+for polymers or protein structures.
+
+A final class of descriptors are fingerprints. Traditionally, these
+descriptors are represented in the form of bit strings. Binary
+fingerprints can be divided into hashed fingerprints, where
+substructures (such as paths of length $n$) are converted to a string
+representation and then hashed into a randomly selected bit positions,
+or keyed fingerprints, where each bit position corresponds to a unique
+substructural feature. The nature of the features can range from
+simple functional groups (hydroxyl, carbonyl) and topological
+substructures (paths, chains, cycles) to atom environments
+\cite{Bremser1978,Bender:2004aa} and pharmacophoric elements
+\cite{Renner:2006aa}.
+
+While the descriptors mentioned so far are derived from some form of
+chemical structure, an alternative approach is to employ observed
+biological activities of a molecule as descriptors of that molecule.
+This approach was taken by Sedykh et al \cite{Sedykh:2011fk}   who
+hypothesized that dose response data points from high throughput
+dose-reponse assays could be employed as biological descriptors. They
+observed that when such descriptors were combined with traditional
+chemical structure descriptors, the predictive performance of models
+developed using the combined set was better than those developed using
+conventional approaches. A similar approach can be taken using the
+PASS methodology \cite{Poroikov:2007aa}, though this is an indirect
+method - one must predict the PASS profile from the chemical
+structure, which can then be used in subsequent predictive models [REF?].
+
+\section{Descriptor Implementations}
+\label{sec:descr-impl}
+
+A variety of descriptor implementations are available, with most
+cheminformatics toolkits providing implementations of a common set of
+descriptors and fingerprints. Table \ref{tab:impl} summarizes the
+descriptor classes implemented by several toolkits and applications.
+
+Given the multiple implementations of many descriptors, it is natural
+to compare their performance. As noted previously, it is desirable,
+but unlikely that implementations of the same descriptor using
+different cheminformatcs toolkits will be identical. Gupta et al
+employed SMARTS based descriptors from the CDK and MOE to develop
+decision tree models to predict human liver microsomal metabolic
+stability\cite{Gupta:2010uq}. Their results indicated very similar
+performance between the two implementations.
+
+One can also directly compare descriptor values generated using
+different implementations.
+
+A number of physicochemical descriptors commonly employed in
+predictive modeling are themselves models of an experimental
+property. For example, log P can be experimentally measured and a
+number of algorithms have been developed to predict log P
+[REFs]. Given the utility of log P in drug discovery, a number of
+implementations are available. Given that this descriptor is a
+surrogate for an actual experimental property, it is reasonable to
+compare calculated log P values from different implementations to the
+experimentall observed values, rather than between themselves. Figure
+\ref{fig:logp} compares computed log P values from the CDK, ChemAxon
+and ACD Labs for a set of 10,000 molecules taken from the logPstar
+dataset [REF].
+
 \section{Running Notes}
 \begin{itemize}
-\item What are descriptors, why are they needed?
+\item \textbf{Done} What are descriptors, why are they needed?
 \item Survey of descriptor types, focusing on nature of inputs,
   complexity, interpretabilty 
-\item Talk about biological descriptors (see paper from Tropsha that
+\item \textbf{Done} Talk about biological descriptors (see paper from Tropsha that
   uses bioactivity data as descriptors)
@@
 
+\ctable[caption={A comparison of Topological Polar Surface Area values
+generated using the CDK and ACD Labs software, for 57,600 molecules
+taken from Pubchem AID 1996},cap={},label={fig:tpsa},botcap,figure]
+{c}
+{}
+{
+  \includegraphics[width=\linewidth]{tpsa-cdk-acd}
+}
+
+\ctable[caption={A comparison of experimental versus calculated logP values
+generated using the CDK, ACD Labs and ChemAxon software, for 10,000 molecules
+taken from the proprietary logPstar dataset [REF?]},cap={},label={fig:logp},botcap,figure]
+{c}
+{}
+{
+  \includegraphics[width=\linewidth]{logp-comp-exptl}
+}
+
 \end{document}

### 2012-03-19 paper.tex
Some minor reformatting and text

@@
 
-A final class of descriptors are fingerprints. Traditionally, these
-descriptors are represented in the form of bit strings. Binary
-fingerprints can be divided into hashed fingerprints, where
-substructures (such as paths of length $n$) are converted to a string
-representation and then hashed into a randomly selected bit positions,
-or keyed fingerprints, where each bit position corresponds to a unique
-substructural feature. The nature of the features can range from
-simple functional groups (hydroxyl, carbonyl) and topological
-substructures (paths, chains, cycles) to atom environments
-\cite{Bremser1978,Bender:2004aa} and pharmacophoric elements
-\cite{Renner:2006aa}.
+\textbf{Fingerprints} A final class of descriptors are
+fingerprints. Traditionally, these descriptors are represented in the
+form of bit strings. Binary fingerprints can be divided into hashed
+fingerprints, where substructures (such as paths of length $n$) are
+converted to a string representation and then hashed into a randomly
+selected bit positions, or keyed fingerprints, where each bit position
+corresponds to a unique substructural feature. The nature of the
+features can range from simple functional groups (hydroxyl, carbonyl)
+and topological substructures (paths, chains, cycles) to atom
+environments \cite{Bremser1978,Bender:2004aa} and pharmacophoric
+elements \cite{Renner:2006aa}. Depending on the definition,
+fingerprints can work with topology only or may require 3D
+conformations. The latter class of fingerprints are usually related to
+3D pharmacophores and most fingerprint definitions only require
+connectivity information.
 
-While the descriptors mentioned so far are derived from some form of
-chemical structure, an alternative approach is to employ observed
-biological activities of a molecule as descriptors of that molecule.
-This approach was taken by Sedykh et al \cite{Sedykh:2011fk}   who
-hypothesized that dose response data points from high throughput
-dose-reponse assays could be employed as biological descriptors. They
-observed that when such descriptors were combined with traditional
-chemical structure descriptors, the predictive performance of models
-developed using the combined set was better than those developed using
-conventional approaches. A similar approach can be taken using the
-PASS methodology \cite{Poroikov:2007aa}, though this is an indirect
-method - one must predict the PASS profile from the chemical
-structure, which can then be used in subsequent predictive models [REF?].
+\textbf{Bioactivity as a descriptor} While the descriptors mentioned
+so far are derived from some form of chemical structure, an
+alternative approach is to employ observed biological activities of a
+molecule as descriptors of that molecule.  This approach was taken by
+Sedykh et al \cite{Sedykh:2011fk} who hypothesized that dose response
+data points from high throughput dose-reponse assays could be employed
+as biological descriptors. They observed that when such descriptors
+were combined with traditional chemical structure descriptors, the
+predictive performance of models developed using the combined set was
+better than those developed using conventional approaches. A similar
+approach can be taken using the PASS methodology
+\cite{Poroikov:2007aa}, though this is an indirect method - one must
+predict the PASS profile from the chemical structure, which can then
+be used in subsequent predictive models [REF?].
 

### 2012-03-20 paper.tex
Added some text around topological descriptors

@@
 for polymers or protein structures.
+
+Broadly, one can classify descriptors based on the nature of
+structural information they require. Constitutional descriptors only
+require atom and bond labels and usually represent counts of different
+types of atoms or bonds. While very simplistic they can play a useful
+role in a variety of applications ranging from summaries of
+physicochemical properties to predictive modeling
+\cite{Bender:2005aa}.
+
+Topological descriptors take into account connectivity along with atom
+and bond labels. These descriptors essentially consider the molecule
+as a labeled graph and characterize it using graph invariants. In
+addition to graph invariants, topological descriptors based on
+information theory such as entropy \cite{Dehmer:2009uq}. An advantage
+of these descriptors is that do not require any intensive
+pre-processing steps such as 3D coordinate generation and
+conformational analysis. In addition, many graph invariants can be
+rapidly computed for graphs corresponding to small molecule structures
+(even though they may be computationally intractable on much larger
+graphs). While this class of descriptors have been used extensively in
+predictive modeling applications [REFS], a key drawback is their lack
+of interpretability. Given the abstract nature of the many graph
+invariants and information theoretic descriptors, their use in
+predictive models makes it difficult to explain the model in simple
+physicochemical terms.
 

### 2012-03-20 paper.tex
Added some more text

@@
 predictive modeling applications [REFS], a key drawback is their lack
-of interpretability. Given the abstract nature of the many graph
+of interpretability. Given the abstract nature of many graph
 invariants and information theoretic descriptors, their use in
@@
 descriptor classes implemented by several toolkits and applications.
+
+While library calls for descriptor evaluation is the most flexible
+approach, it is not useful for non-programmers as well as users who
+simply need a descriptor matrix for modeling or searching. As a
+result, GUI or command line wrappers around descriptor APIs enable
+easier usage. The CDKDescUI
+application(\url{http://rguha.net/code/java/cdkdesc.html}) is a Java
+Swing application that exposes the CDK descriptor and fingerprint
+API's. It allows the users to specify a SMILES or SD file and evaluate
+a selected set of descriptors or fingerprints and export them to a
+text file. The application also has a command line mode that can be
+run in server environments. A similar application is PaDEL
+(\url{http://padel.nus.edu.sg/software/padeldescriptor/}) that
+includes the CDK descriptors and also implements a number of
+additional descriptors [SOME MENTION OF THESE?]
+
+In addition to libraries and their wrappers there are various GUI
+applications that allow easy descriptor calculations. Examples include
+Bioclipse \cite{Spjuth:2007aa}, MOE (Chemical Computing Group) and
+Maestro (Schr\"{o}dinger, Inc.)
+
+\subsection{Comparing implementations}
+\label{sec:comp-impl}
 

### 2012-04-01 paper.tex
Added some more text
Added more text about descriptor categories

@@
 graphs). While this class of descriptors have been used extensively in
-predictive modeling applications [REFS], a key drawback is their lack
-of interpretability. Given the abstract nature of many graph
-invariants and information theoretic descriptors, their use in
-predictive models makes it difficult to explain the model in simple
-physicochemical terms.
+predictive modeling applications
+\cite{Garcia-Domenech:2008aa,Randic:2001ac,Randic:2001ad,Besalu:2001aa,jcics:1995:35:272,Kier:1986ae},
+a key drawback is their lack of interpretability. Given the abstract
+nature of many graph invariants and information theoretic descriptors,
+their use in predictive models makes it difficult to explain the model
+in simple physicochemical terms, though various attempts have been  
+made at interpretations \cite{Todeschini:1975dq,Stanton:2003aa}.
 
-\textbf{Fingerprints} A final class of descriptors are
-fingerprints. Traditionally, these descriptors are represented in the
-form of bit strings. Binary fingerprints can be divided into hashed
-fingerprints, where substructures (such as paths of length $n$) are
-converted to a string representation and then hashed into a randomly
-selected bit positions, or keyed fingerprints, where each bit position
-corresponds to a unique substructural feature. The nature of the
-features can range from simple functional groups (hydroxyl, carbonyl)
-and topological substructures (paths, chains, cycles) to atom
-environments \cite{Bremser1978,Bender:2004aa} and pharmacophoric
-elements \cite{Renner:2006aa}. Depending on the definition,
-fingerprints can work with topology only or may require 3D
+Geometric descriptors require a 3D conformation as input and therefore
+are not as fast as topological descriptors which do not require the 3D
+coordinate generation step. Nearly all geometric descriptors work on a
+single conformation and for cases where multiple conformation must be
+considered, an averaging procedure is usually employed.
+
+A number of geometric descriptors characterize the molecular
+surface. The simplest such example is the surface area descriptor (and
+the corresponding volume descriptor). A simple extension of such a
+descriptor is to characterize distributions of a physicochemical
+property over the molecular surface. Note that surface descriptor
+calculations based onanalytical \cite{Connolly:1983aa} or tesselation
+algorithms \cite{Eisenhaber:1995qf} can be slow and that parametrized
+methods such as the VSA descriptors \cite{Labute:2008aa} that are
+based on precalculated surface area values derived from a list of
+functional groups. A comparison of surface area values generated by
+the two approaches for XXX molecules indicates a high degree of
+correlation.
+
+\textbf{Fingerprints} Traditionally, fingerprint descriptors are
+repreented in the form of bit strings. Binary fingerprints can be
+divided into hashed fingerprints, where substructures (such as paths
+of length $n$) are converted to a string representation and then
+hashed into a randomly selected bit positions, or keyed fingerprints,
+where each bit position corresponds to a unique substructural
+feature. The nature of the features can range from simple functional
+groups (hydroxyl, carbonyl) and topological substructures (paths,
+chains, cycles) to atom environments \cite{Bremser1978,Bender:2004aa}
+and pharmacophoric elements \cite{Renner:2006aa}. Depending on the
+definition, fingerprints can work with topology only or may require 3D
 conformations. The latter class of fingerprints are usually related to
@@
 connectivity information.
+
+Fingerprints are a compact representation of a chemical structure and
+are used in a variety of ways ranging from database searches looking
+for similar structures or substructures 
+to virtual screenings studies
+\cite{Sun:2006kx}. In many search applications, the binary nature of
+fingerprints allow for efficient algorithms to compare fingerprints
+and a number of heuristics have been developed that allow for rapid
+similarity searches \cite{Baldi:2008aa,Swamidass:2007ab}.
+
+\textbf{Descriptors for inorganic compounds} Inorganic materials are
+sufficiently different from small molecules that traditional
+descriptors are not very useful. Recently, a number of descriptors for
+inorganic materials are very similar to constitutional descriptors -
+essentially counts of various crystal parameters such as acccessible
+surface areas and volumes
+\cite{Willems:2012zr,Haranczyk:2010ys,Mackay:1984ve}. Other
+descriptors characterize the periodic voids of porous materials via
+tesselation methods \cite{Anurova:2009ly,Carr:2009kx}
 
@@
 predict the PASS profile from the chemical structure, which can then
-be used in subsequent predictive models [REF?].
+be used in subsequent predictive models.
 
@@
 result, GUI or command line wrappers around descriptor APIs enable
-easier usage. The CDKDescUI
-application(\url{http://rguha.net/code/java/cdkdesc.html}) is a Java
-Swing application that exposes the CDK descriptor and fingerprint
-API's. It allows the users to specify a SMILES or SD file and evaluate
-a selected set of descriptors or fingerprints and export them to a
-text file. The application also has a command line mode that can be
-run in server environments. A similar application is PaDEL
-(\url{http://padel.nus.edu.sg/software/padeldescriptor/}) that
-includes the CDK descriptors and also implements a number of
-additional descriptors [SOME MENTION OF THESE?]
+easier usage. The CDKDescUI application
+(\url{http://rguha.net/code/java/cdkdesc.html}) is a Java Swing
+application that exposes the CDK descriptor and fingerprint API's. It
+allows the users to specify a SMILES or SD file and evaluate a
+selected set of descriptors or fingerprints and export them to a text
+file. The application also has a command line mode that can be run in
+server environments. A similar application is PaDEL \cite{Yap:2011fk}
+that includes the CDK descriptors and also implements a number of
+additional descriptors such as the extended topochemical atom
+descriptors \cite{Roy:2004uq}, McGowan Volume \cite{Zhao:2003kx}, and
+the Klekota \& Roth substructure counts \cite{Klekota:2008vn}.
 
@@
 Bioclipse \cite{Spjuth:2007aa}, MOE (Chemical Computing Group) and
-Maestro (Schr\"{o}dinger, Inc.)
+Maestro (Schr\"{o}dinger, Inc.). Workflow tools are also a useful
+class of applications for descriptor generation and allow one to
+easily generate descriptors from multiple sources.
 

### 2012-04-01 paper.tex
Added more text ondescriptor types

@@
 descriptor is to characterize distributions of a physicochemical
-property over the molecular surface. Note that surface descriptor
-calculations based onanalytical \cite{Connolly:1983aa} or tesselation
-algorithms \cite{Eisenhaber:1995qf} can be slow and that parametrized
-methods such as the VSA descriptors \cite{Labute:2008aa} that are
-based on precalculated surface area values derived from a list of
-functional groups. A comparison of surface area values generated by
-the two approaches for XXX molecules indicates a high degree of
-correlation.
+property over the molecular surface such as the Charged Partial
+Surface Area descriptors \cite{Stanton:1990aa} that characterize the
+partial charge distribution over the molecular surfce. Note that
+surface descriptor calculations based on analytical
+\cite{Connolly:1983aa} or tesselation algorithms
+\cite{Eisenhaber:1995qf} can be slow and that parametrized methods
+such as the VSA descriptors \cite{Labute:2008aa} that are based on
+precalculated surface area values derived from a list of functional
+groups. A comparison of surface area values generated by the two
+approaches for XXX molecules indicates a high degree of correlation.
+\textbf{Ray tracing descriptors, ultrafast shape descriptors,
+  spherical harmonic shape descriptors, field based descriptors
+  [Cresset etc]}
 
@@
 are used in a variety of ways ranging from database searches looking
-for similar structures or substructures 
-to virtual screenings studies
+for similar structures or substructures to virtual screenings studies
 \cite{Sun:2006kx}. In many search applications, the binary nature of
-fingerprints allow for efficient algorithms to compare fingerprints
-and a number of heuristics have been developed that allow for rapid
-similarity searches \cite{Baldi:2008aa,Swamidass:2007ab}.
+fingerprints allow for efficient algorithms to compare molecules and
+evaluate similarity and a number of heuristics have been developed
+that allow for rapid similarity searches
+\cite{Baldi:2008aa,Swamidass:2007ab}. In virtual screening scenarios,
+fingerprints can be used as features. In many cases, the entire
+fingerprint is employed as a set of independent variables. Since this
+can lead to poor predictive models, some sort of feature selection is
+usually employed and there have been studies that address the problem
+of identifying subsets of bits from a fingerprint that lead to optimal
+performance in predictive models (albeit, usually in a target-class
+specific manner) \cite{Xue:2004fk}.
 
@@
   uses bioactivity data as descriptors)
-\item Discuss how fingerprints can also be used as descriptors
+\item \textbf{Done} Discuss how fingerprints can also be used as descriptors
 \item Also mention descriptors for non-small molecule cases
@@
   \item proteins
-  \item polymers
-  \item inorganics (maybe ask Maciej for pointers)
+  \item polymers \& mixtures
+  \item \textbf{Done} inorganics (maybe ask Maciej for pointers)
   \end{itemize}

### 2012-04-08 paper.tex
Added some more text on shape descriptors

@@
 groups. A comparison of surface area values generated by the two
-approaches for XXX molecules indicates a high degree of correlation.
-\textbf{Ray tracing descriptors, ultrafast shape descriptors,
+approaches for XXX molecules indicates a high degree of
+correlation.
+
+While molecular surfaces are useful, molecular shape is a key factor
+in ligand-receptor interactions. A variety of descriptors have been
+devised to characterize molecular shape. For example, the ``shape
+signature'' approach described by Zauhar et al \cite{Zauhar:2003fk} is a novel application of a
+ray tracing method that generates a description of the molecular
+shape. Ballester et al \cite{Ballester:2007aa} described the Ultrafast
+Shape Recognition (USR) approach that characterized molecular shape in
+terms of the distribution of distances between the atoms and a set of
+pre-defined points. As a result, the calculation procedure is
+extremely fast and once calculated shape similarities can be evaluated
+rapidly via a simple distance calculation. 
+\textbf{
   spherical harmonic shape descriptors, field based descriptors

### 2012-04-08 paper.tex
More text of descriptors of shape

@@
 in ligand-receptor interactions. A variety of descriptors have been
-devised to characterize molecular shape. For example, the ``shape
-signature'' approach described by Zauhar et al \cite{Zauhar:2003fk} is a novel application of a
-ray tracing method that generates a description of the molecular
-shape. Ballester et al \cite{Ballester:2007aa} described the Ultrafast
-Shape Recognition (USR) approach that characterized molecular shape in
-terms of the distribution of distances between the atoms and a set of
-pre-defined points. As a result, the calculation procedure is
-extremely fast and once calculated shape similarities can be evaluated
-rapidly via a simple distance calculation. 
-\textbf{
-  spherical harmonic shape descriptors, field based descriptors
-  [Cresset etc]}
+devised to characterize molecular shape. The use of overlapping atom
+centered Gaussians to represent molecular shape \cite{Grant:1995aa}
+represents a significant computational advantage over the traditional
+analytical approach of intersecting hard spheres, supporting extremely
+rapid shape comparisons \cite{Grant:1999vn}. An alternative approach
+described by Ritchie et al \cite{Ritchie:1999kx} employed a spherical
+harmonic based representation of surfaces, that in contrast to the
+Gaussian approach allows one to generate characterizations of the
+surface and hence the shape at varying resolutions. In addition to
+functional representations of molecular shape, a number of simpler
+formulations have been defined. For example, the ``shape signature''
+approach described by Zauhar et al \cite{Zauhar:2003fk} is a novel
+application of a ray tracing method that generates a description of
+the molecular shape. Ballester et al \cite{Ballester:2007aa} described
+the Ultrafast Shape Recognition (USR) approach that characterized
+molecular shape in terms of the distribution of distances between the
+atoms and a set of pre-defined points. As a result, the calculation
+procedure is extremely fast and once calculated shape similarities can
+be evaluated rapidly via a simple distance calculation.
 

### 2012-04-08 paper.tex
More etext of shape descriptors

@@
 Gaussian approach allows one to generate characterizations of the
-surface and hence the shape at varying resolutions. In addition to
-functional representations of molecular shape, a number of simpler
-formulations have been defined. For example, the ``shape signature''
-approach described by Zauhar et al \cite{Zauhar:2003fk} is a novel
-application of a ray tracing method that generates a description of
-the molecular shape. Ballester et al \cite{Ballester:2007aa} described
-the Ultrafast Shape Recognition (USR) approach that characterized
-molecular shape in terms of the distribution of distances between the
-atoms and a set of pre-defined points. As a result, the calculation
+surface and hence the shape at varying resolutions. Another approach
+to multi-resolution shape representations is the use of $\alpha$
+shapes \cite{Edelsbrunner:1994aa}, which represent a parametrization
+of the convex hull derived from the molecular coordinates. This
+approach has been employed to characterize shape and volume
+\cite{Wilson:2009ys} though it is more computationally intensive than
+the methods mentioned previously. In addition to functional
+representations of molecular shape, a number of simpler formulations
+have been defined. For example, the ``shape signature'' approach
+described by Zauhar et al \cite{Zauhar:2003fk} is a novel application
+of a ray tracing method that generates a description of the molecular
+shape. Ballester et al \cite{Ballester:2007aa} described the Ultrafast
+Shape Recognition (USR) approach that characterized molecular shape in
+terms of the distribution of distances between the atoms and a set of
+pre-defined points. This procedure characterizes the shape in the form
+of a 12-D real valued fingerprint. As a result, the calculation
 procedure is extremely fast and once calculated shape similarities can
@@
 specific manner) \cite{Xue:2004fk}.
+
+Finally, quantum mechanical calculations can also generated
+descriptors that can be applied to problems such as QSAR models
+(though of course the time complexity of QM derived descriptors do not
+make them amenable for large scale virtual screening applications).
 

### 2012-04-08 paper.tex
Started on polymer descriptors

@@
 
+Finally, quantum mechanical calculations can also generated
+descriptors that can be applied to problems such as QSAR models
+(though of course the time complexity of QM derived descriptors do not
+make them amenable for large scale virtual screening applications).
+\textbf{QM descriptors}
+
+\textbf{Descriptors for polymeric systems} Mention the AA suite of
+descriptors to characterize things like proteins. Consider related
+protein descriptors Mention DNA
+descriptors. Mention the polymer descriptors and try and find soem
+more.
+
+Polymeric systems represent a challenge for descriptor development. A
+number of approaches have been described that includes both
+deterministic, geometry derived descriptors \cite{Edvinsson:2003bh} as
+well as stochastic methods \cite{Gonzales-Diaz:2003ly}.
+
 \textbf{Fingerprints} Traditionally, fingerprint descriptors are
@@
 specific manner) \cite{Xue:2004fk}.
-
-Finally, quantum mechanical calculations can also generated
-descriptors that can be applied to problems such as QSAR models
-(though of course the time complexity of QM derived descriptors do not
-make them amenable for large scale virtual screening applications).
 

### 2012-04-09 paper.tex
Started on polymer descriptors
Added some more text on qm descs

@@
 single conformation and for cases where multiple conformation must be
-considered, an averaging procedure is usually employed.
+considered, an averaging procedure is usually employed. Examples of
+these descriptors include the gravitational index [REF] and moment of
+inertia descriptors. However, geometric descriptors encompass many
+aspects of a molecular structure. For example, a number of geometric
+descriptors characterize the molecular surface. The simplest such
+example is the surface area descriptor (and the corresponding volume
+descriptor). A simple extension of such a descriptor is to
+characterize distributions of a physicochemical property over the
+molecular surface such as the Charged Partial Surface Area descriptors
+\cite{Stanton:1990aa} that characterize the partial charge
+distribution over the molecular surfce. Note that surface descriptor
+calculations based on analytical \cite{Connolly:1983aa} or tesselation
+algorithms \cite{Eisenhaber:1995qf} can be slow and that parametrized
+methods such as the VSA descriptors \cite{Labute:2008aa} that are
+based on precalculated surface area values derived from a list of
+functional groups. A comparison of surface area values generated by
+the the tesselatin method and the VSA method (as implemented in the
+CDK) for XXX molecules indicates a high degree of correlation.
 
-A number of geometric descriptors characterize the molecular
-surface. The simplest such example is the surface area descriptor (and
-the corresponding volume descriptor). A simple extension of such a
-descriptor is to characterize distributions of a physicochemical
-property over the molecular surface such as the Charged Partial
-Surface Area descriptors \cite{Stanton:1990aa} that characterize the
-partial charge distribution over the molecular surfce. Note that
-surface descriptor calculations based on analytical
-\cite{Connolly:1983aa} or tesselation algorithms
-\cite{Eisenhaber:1995qf} can be slow and that parametrized methods
-such as the VSA descriptors \cite{Labute:2008aa} that are based on
-precalculated surface area values derived from a list of functional
-groups. A comparison of surface area values generated by the two
-approaches for XXX molecules indicates a high degree of
-correlation.
-
-While molecular surfaces are useful, molecular shape is a key factor
-in ligand-receptor interactions. A variety of descriptors have been
-devised to characterize molecular shape. The use of overlapping atom
-centered Gaussians to represent molecular shape \cite{Grant:1995aa}
-represents a significant computational advantage over the traditional
-analytical approach of intersecting hard spheres, supporting extremely
-rapid shape comparisons \cite{Grant:1999vn}. An alternative approach
-described by Ritchie et al \cite{Ritchie:1999kx} employed a spherical
-harmonic based representation of surfaces, that in contrast to the
-Gaussian approach allows one to generate characterizations of the
-surface and hence the shape at varying resolutions. Another approach
-to multi-resolution shape representations is the use of $\alpha$
-shapes \cite{Edelsbrunner:1994aa}, which represent a parametrization
-of the convex hull derived from the molecular coordinates. This
-approach has been employed to characterize shape and volume
-\cite{Wilson:2009ys} though it is more computationally intensive than
-the methods mentioned previously. In addition to functional
-representations of molecular shape, a number of simpler formulations
-have been defined. For example, the ``shape signature'' approach
-described by Zauhar et al \cite{Zauhar:2003fk} is a novel application
-of a ray tracing method that generates a description of the molecular
-shape. Ballester et al \cite{Ballester:2007aa} described the Ultrafast
-Shape Recognition (USR) approach that characterized molecular shape in
-terms of the distribution of distances between the atoms and a set of
-pre-defined points. This procedure characterizes the shape in the form
-of a 12-D real valued fingerprint. As a result, the calculation
-procedure is extremely fast and once calculated shape similarities can
-be evaluated rapidly via a simple distance calculation.
+Molecular shape is a key factor in ligand-receptor interactions and is
+an important approach to virtual screening. A variety of descriptors
+have been devised to characterize molecular shape. The use of
+overlapping atom centered Gaussians to represent molecular shape
+\cite{Grant:1995aa} represents a significant computational advantage
+over the traditional analytical approach of intersecting hard spheres,
+supporting extremely rapid shape comparisons \cite{Grant:1999vn}. An
+alternative approach described by Ritchie et al \cite{Ritchie:1999kx}
+employed a spherical harmonic based representation of surfaces, that
+in contrast to the Gaussian approach allows one to generate
+characterizations of the surface and hence the shape at varying
+resolutions. Another approach to multi-resolution shape
+representations is the use of $\alpha$ shapes
+\cite{Edelsbrunner:1994aa}, which represent a parametrization of the
+convex hull derived from the molecular coordinates. This approach has
+been employed to characterize shape and volume \cite{Wilson:2009ys}
+though it is more computationally intensive than the methods mentioned
+previously. In addition to functional representations of molecular
+shape, a number of simpler formulations have been defined. For
+example, the ``shape signature'' approach described by Zauhar et al
+\cite{Zauhar:2003fk} is a novel application of a ray tracing method
+that generates a description of the molecular shape. Ballester et al
+\cite{Ballester:2007aa} described the Ultrafast Shape Recognition
+(USR) approach that characterized molecular shape in terms of the
+distribution of distances between the atoms and a set of pre-defined
+points. This procedure characterizes the shape in the form of a 12-D
+real valued fingerprint. As a result, the calculation procedure is
+rapid and once calculated shape similarities can be evaluated rapidly
+via a simple distance calculation.
 
@@
 (though of course the time complexity of QM derived descriptors do not
-make them amenable for large scale virtual screening applications).
-\textbf{QM descriptors}
+make them amenable for large scale virtual screening
+applications). These descriptors can be evaluated at various levels of
+theory. For example, partial charges, HOMO and LUMO energies are
+commonly evaluated using semi-empirical methods, while Eroglu and
+T\"{u}rkmen \cite{Eroglu:2007uq} described DFT-derived descriptors
+such as electronegativity, hardness and ionization potential (though
+Puzyn et al question the need for such a high level of theory
+\cite{Puzyn:2008fk}).
+
+We next consider descriptors that are designed for specific classes of
+molecular systems.
 
@@
 
-Polymeric systems represent a challenge for descriptor development. A
+Polymeric systems represent a challenge for descriptor
+development. There have been reports that employed tradtional
+small-molecule descriptors \cite{Mattioni:2002dq,Katritzky:1998cr}
+that consider the polymer in terms of the individual monomers. A
 number of approaches have been described that includes both

### 2012-04-09 paper.tex
more or less done on descriptor categorization

@@
 considered, an averaging procedure is usually employed. Examples of
-these descriptors include the gravitational index [REF] and moment of
-inertia descriptors. However, geometric descriptors encompass many
-aspects of a molecular structure. For example, a number of geometric
-descriptors characterize the molecular surface. The simplest such
-example is the surface area descriptor (and the corresponding volume
-descriptor). A simple extension of such a descriptor is to
-characterize distributions of a physicochemical property over the
-molecular surface such as the Charged Partial Surface Area descriptors
-\cite{Stanton:1990aa} that characterize the partial charge
-distribution over the molecular surfce. Note that surface descriptor
-calculations based on analytical \cite{Connolly:1983aa} or tesselation
-algorithms \cite{Eisenhaber:1995qf} can be slow and that parametrized
-methods such as the VSA descriptors \cite{Labute:2008aa} that are
-based on precalculated surface area values derived from a list of
-functional groups. A comparison of surface area values generated by
-the the tesselatin method and the VSA method (as implemented in the
-CDK) for XXX molecules indicates a high degree of correlation.
+these descriptors include the gravitational index
+\cite{Katritzky:1996ly} and moment of inertia descriptors. However,
+geometric descriptors encompass many aspects of a molecular
+structure. For example, a number of geometric descriptors characterize
+the molecular surface. The simplest such example is the surface area
+descriptor (and the corresponding volume descriptor). A simple
+extension of such a descriptor is to characterize distributions of a
+physicochemical property over the molecular surface such as the
+Charged Partial Surface Area descriptors \cite{Stanton:1990aa} that
+characterize the partial charge distribution over the molecular
+surfce. Note that surface descriptor calculations based on analytical
+\cite{Connolly:1983aa} or tesselation algorithms
+\cite{Eisenhaber:1995qf} can be slow and that parametrized methods
+such as the Topological Polar Surface Area (TPSA) \cite{Ertl:2000aa}
+and VSA descriptors \cite{Labute:2008aa} that are based on
+precalculated surface area values derived from a list of functional
+groups. A comparison of surface area values generated by the the
+tesselatin method and the VSA method (as implemented in the CDK) for
+XXX molecules indicates a high degree of correlation.
 
@@
 rapid and once calculated shape similarities can be evaluated rapidly
-via a simple distance calculation.
+via a simple distance calculation. It should also be noted that shape
+indices based on the molecular topology have also been described
+\cite{Kier:1985aa}, which are very rapid to calculate (somewhat
+analogous to the VSA descriptors) but do not directly represent the
+molecular shape and instead
 
@@
 
-We next consider descriptors that are designed for specific classes of
-molecular systems.
-
-\textbf{Descriptors for polymeric systems} Mention the AA suite of
-descriptors to characterize things like proteins. Consider related
-protein descriptors Mention DNA
-descriptors. Mention the polymer descriptors and try and find soem
-more.
-
-Polymeric systems represent a challenge for descriptor
-development. There have been reports that employed tradtional
-small-molecule descriptors \cite{Mattioni:2002dq,Katritzky:1998cr}
-that consider the polymer in terms of the individual monomers. A
-number of approaches have been described that includes both
-deterministic, geometry derived descriptors \cite{Edvinsson:2003bh} as
-well as stochastic methods \cite{Gonzales-Diaz:2003ly}.
+The descriptor categories listed here are not necessarily distinct,
+many hybrid descriptors have been described that combine one or more
+categories. For example, the charge indices described by Galvez et al
+\cite{Galvez:1994aa} combine partial charge information and
+topological connectivity. Similarly the CPSA descriptors combine
+surface area and partial charge information. When considering hybrid
+descriptors, it is useful to note the autocorrelation descriptors that
+characterize the distribution of molecular properties across a
+molecule in terms of an autocorrelation function. These descriptors
+may be based on topological or geometric distances and can be combined
+with arbitrary molecular properties \cite{Moreau:1980aa,Broto:1984aa}.
 
 \textbf{Fingerprints} Traditionally, fingerprint descriptors are
-repreented in the form of bit strings. Binary fingerprints can be
+represented in the form of bit strings. Binary fingerprints can be
 divided into hashed fingerprints, where substructures (such as paths
@@
 
+We next consider descriptors that are designed for specific classes of
+molecular systems.
+
+\textbf{Descriptors for polymeric systems} Polymeric systems represent
+a challenge for descriptor development. There have been reports that
+employed tradtional small-molecule descriptors
+\cite{Mattioni:2002dq,Katritzky:1998cr} that consider the polymer in
+terms of the individual monomers. A number of approaches have been
+described that includes both deterministic, geometry derived
+descriptors \cite{Edvinsson:2003bh} as well as stochastic methods
+\cite{Gonzales-Diaz:2003ly}. For the case of polypeptides the
+descriptors generally tend to be based on the individual amino
+acids. Examples include the VHSE family of descriptors described by
+Mei et al \cite{Mei:2005kx}, a set of side chain descriptors from
+Collantes and Dunn \cite{Collantes:1995vn}, field based descriptors
+\cite{Norinder:1991ys} and the AAindex database
+\cite{Kawashima:1999zr} which allows rapid lookup of amino acid
+properties.
+
 \textbf{Descriptors for inorganic compounds} Inorganic materials are
 sufficiently different from small molecules that traditional
-descriptors are not very useful. Recently, a number of descriptors for
+descriptors are not very useful. Recently, a number of descriptors for
 inorganic materials are very similar to constitutional descriptors -

### 2012-04-09 paper.tex
started section on versioning

@@
 sufficiently different from small molecules that traditional
-descriptors are not very useful. Recently, a number of descriptors for
+descriptors are not very useful. Recently, a number of descriptors for
 inorganic materials are very similar to constitutional descriptors -
@@
 
+\subsection{Descriptor Versioning}
+\label{sec:descr-vers}
+
+As noted above, different implementations of the same descriptor can
+generate slightly different values depending on the underlying
+chemistry model, differences in reference values and so on. Even
+within the same software implementation, different versions of the
+same descriptor can generate different values. Thus to maintain
+reproducibility in sudies that employ molecular descriptors, some form
+of versioning is crucial. The simplest form of version is to manually
+note the version of the implementation used for the calculation and
+report it along with the results. Though useful, it is not amenable to
+automation. In some cases, toolkits will provide an API call to
+retrieve the version information, which can then be associated with a
+descriptor calculation.
+
 \section{Running Notes}

### 2012-04-09 paper.tex
Done with desc versioning section

@@
 
-\subsection{Descriptor Versioning}
+\subsection{Descriptor Naming \& Versioning}
 \label{sec:descr-vers}
@@
 descriptor calculation.
+
+A topic closely related to versioning is descriptor naming. The
+preceding sections have discussed a categorization of descriptors and
+have used a number of commonly used names. When performing
+calculations in an automated manner, or employing remote services for
+descriptor calculations \cite{Dong:2007aa}, it is useful to be able to
+refer to descriptors via a standardized naming scheme. Currently the
+QSAR modeling community lacks such a scheme, though the Chemistry
+Developmen Kit (CDK), a Java library for cheminformatics
+\cite{Steinbeck:2006aa} does implement such a scheme. The ``descriptor
+specification'' approach defines a set of classes that include
+information of the implementation title, identifier and vendor. The
+identifier may or may not incude a version number. The actual details
+of the a descriptor are listed in a dictionary, along with an optional
+namespace. The dictionary entry is referenced by the descriptor
+specification and this design thus allows one to seamlessly work with
+multiple implementations of the same descriptor from different vendors
+and keep track of which version is employed throughout a calculation
+or workflow. In addition, the CDK provides a hierarchical annotation
+of descriptors, grouping them into different classes. Since this
+annotation is based on RDF \cite{Taylor:2006ab}, it allows one to go
+beyond simple groupings and perform reasoning on descriptors. A simple
+example is to identify functionally equivalent descriptors (such as
+simple transformations of a base descriptor - square and cube roots of
+the gravitational index is an example \cite{Wessel:1998ve}) in an
+automated fashion. While this infrastructure is currently only
+supported by the CDK, adoption of it (or some similar framework) would
+enable easier interoperability between descriptor implementations and
+enhance reproducibility.
 

### 2012-04-09 paper.tex
Started adding impl table

@@
 
+Given the focus of this article on Open Source implementations, we
+have noted the availability in Table \ref{tab:impl}. While a number of
+implementations are not strictly Open Source according to OSI
+definitions, the fact that they provide free academic licenses does
+allow them to be used somewhat freely.
+
 \subsection{Comparing implementations}
@@
 to compare their performance. As noted previously, it is desirable,
-but unlikely that implementations of the same descriptor using
+but unlikely, that implementations of the same descriptor using
 different cheminformatcs toolkits will be identical. Gupta et al
@@
 property. For example, log P can be experimentally measured and a
-number of algorithms have been developed to predict log P
-[REFs]. Given the utility of log P in drug discovery, a number of
-implementations are available. Given that this descriptor is a
-surrogate for an actual experimental property, it is reasonable to
-compare calculated log P values from different implementations to the
-experimentall observed values, rather than between themselves. Figure
-\ref{fig:logp} compares computed log P values from the CDK, ChemAxon
-and ACD Labs for a set of 10,000 molecules taken from the logPstar
-dataset [REF].
+number of algorithms have been developed to predict $\log P$
+\cite{Wang:1997aa,Wildman:1999aa}. Given the utility of log P in drug
+discovery, a number of implementations are available. Given that this
+descriptor is a surrogate for an actual experimental property, it is
+reasonable to compare calculated log P values from different
+implementations to the experimentally observed values, rather than
+between themselves. Figure \ref{fig:logp} compares computed log P
+values from the CDK, ChemAxon and ACD Labs for a set of 10,000
+molecules taken from the logPstar dataset [REF].
 
@@
 \begin{itemize}
-\item \textbf{Done} What are descriptors, why are they needed?
-\item Survey of descriptor types, focusing on nature of inputs,
-  complexity, interpretabilty 
-\item \textbf{Done} Talk about biological descriptors (see paper from Tropsha that
-  uses bioactivity data as descriptors)
-\item \textbf{Done} Discuss how fingerprints can also be used as descriptors
-\item Also mention descriptors for non-small molecule cases
-  \begin{itemize}
-  \item proteins
-  \item polymers \& mixtures
-  \item \textbf{Done} inorganics (maybe ask Maciej for pointers)
-  \end{itemize}
 \item Survey of descriptor implementations (OSS and non-OSS)
@@
 
+%% See http://code.google.com/p/rdkit/w/list for RDKit
+\ctable[caption={A summary of descriptor implementations covering Open
+  Source and commercial cases.}, cap={}, label={tab:impl}]
+{llllp{0.3\linewidth}p{0.2\linewidth}}
+{}
+{
+\FL
+Name & License & Toolkit & Application & Descriptors & Fingerprints
+\ML
+CDK & LGPL & Yes & \raggedright{CDKDescUI} &
+\raggedright{Constitutional, topological, geometric, electronic and hybrid} &
+  \raggedright{MACCS, PubChem, Kier \& Hall Estate keys. Hashed linear
+  and circular} \NN
+JChem & Commercial & Yes & cxcalc & \raggedright{} & raggedright{} \NN
+JOELib & GPL & Yes & & \raggedright{} & \raggedright{} \NN
+OpenBabel & GPL & Yes & & \raggedright{} & \raggedright{} \NN
+RDKit & New BSD & Yes & & \raggedright{Constitutional, topological,
+  EState, partial charges and VSA combining EState, partial charges
+  and MR} & MACCS, hashed, atom pairs, topological
+torsions 
+\LL
+}
+
 \ctable[caption={A comparison of Topological Polar Surface Area values

### 2012-04-09 paper.tex
Put in abstract

@@
 \begin{abstract}
-Bar
+  Numerical characterization of molecular structure is a first step in
+  every computational analysis of molecular data. These numerical
+  representations, termed descriptors, come in many forms, ranging
+  from simple atom counts and invariants of the molecular graph to
+  distribution of properties such as charge across a molecular
+  surface. In this article we first present a broad categorization of
+  descriptors and then describe applications and toolkits that can be
+  employed to evaluate them. We highight a number of issues
+  surrounding molecular descriptor calculations such as versioning and
+  reproducibility and describe how some toolkits have attempted to
+  address these problems.
 \end{abstract}
@@
 descriptors and fingerprints. Table \ref{tab:impl} summarizes the
-descriptor classes implemented by several toolkits and applications.
+descriptor classes implemented by several toolkits and
+applications. Note that that due to space limitations this is not a
+comprehensive list of implemented descriptors across different
+applications and the reader is referred to the documentation
+accompanying the software packages.
 
@@
 
+\section{Summary}
+\label{sec:summary}
+
 \section{Running Notes}
@@
   \item MOE
-  \item DRAGON
-  \item CDK
-  \item RDKit
-  \item JChem
-  \item PipelinePilot
   \item OEChem/GraphTK
-  \item JOELib
-  \item CODESSA
-  \item Schrodinger stuff (Meastro?)
   \end{itemize}
@@
   \end{itemize}
-\item Compare OSS descs to commercial ones - point to the Pfizer/CDD
-  paper that compared MOE and CDK
 \end{itemize}
@@
 %% See http://code.google.com/p/rdkit/w/list for RDKit
+%% See
+%% http://www.ra.cs.uni-tuebingen.de/software/joelib/tutorial/descriptors/fingerprints.html for JOELib
+%% http://www.ra.cs.uni-tuebingen.de/software/joelib/tutorial/applications/descriptorcalculation.html for descs
 \ctable[caption={A summary of descriptor implementations covering Open
@@
 \ML
-CDK & LGPL & Yes & \raggedright{CDKDescUI} &
+CDK & LGPL & Yes & Yes &
 \raggedright{Constitutional, topological, geometric, electronic and hybrid} &
@@
   and circular} \NN
-JChem & Commercial & Yes & cxcalc & \raggedright{} & raggedright{} \NN
-JOELib & GPL & Yes & & \raggedright{} & \raggedright{} \NN
+CODESSA & Commercial & No & Yes & \raggedright{Constitutional,
+  topological, geometric, electronic, QM, thermodynamic} & \NN
+Dragon & Commercial & No & Yes & \raggedright{Extensive list
+of all classes of descriptor types and implements all those described
+in Ref.~\citen{Todeschini:2002aa}} &  \NN
+JChem & Commercial & Yes & Yes & \raggedright{Consititutional,
+  geometric, topological, electronic, hybrid} & \raggedright{Hashed,
+  circular, ECFP/FCFP} \NN
+JOELib & GPL & Yes & Yes & \raggedright{} &
+\raggedright{Keys from Xue et al\ \cite{Godden:2000qf}} \NN
 OpenBabel & GPL & Yes & & \raggedright{} & \raggedright{} \NN

### 2012-04-17 paper.tex
Added some more text on desc imp;

@@
   from simple atom counts and invariants of the molecular graph to
-  distribution of properties such as charge across a molecular
+  distribution of properties, such as charge, across a molecular
   surface. In this article we first present a broad categorization of
@@
 large collection of them, approaches to identify a suitable subset
-have been discussed extensively in the literature [REFS].
+have been discussed extensively in the literature
+\cite{Miller:2002aa,Kohavi:1997gf}.
 
@@
 must one choose one or more descriptors that are relevant to the
-problem at hand, but in addition, one must be concerned about how they
+problem at hand, but one must be concerned about how they
 are calculated and whether such a calculation can be reproduced across
@@
 precalculated surface area values derived from a list of functional
-groups. A comparison of surface area values generated by the the
-tesselatin method and the VSA method (as implemented in the CDK) for
-XXX molecules indicates a high degree of correlation.
+groups.  Molecular volume is also a useful physical descriptor and
+both analytical \cite{Eisenhaber:1995qf} and approximate methods are
+available to evaluate it. A very fast method is the VABC approach
+\cite{Zhao:2003fk} that is based on group contributions. It has been
+implemented in the CDK and correlates well with other approximations
+to the molecular volume (Figure \ref{fig:molvol}).
 
@@
 analogous to the VSA descriptors) but do not directly represent the
-molecular shape and instead
+molecular shape and instead infer it from topological connectivity.
 
@@
 
+Due to our experience with the CDK project we provide a brief overview
+of the architecture of the descriptor calculations implemented by the
+CDK. First, the toolkit explicitly differentiates between atom, bond
+and whole molecule descriptors. There are a total of 30, 6 and 51
+descriptors for atoms, bonds and molecules respectively. Note that
+each individual descriptor may calculate multiple values. For example,
+the CPSA descriptor evaluates 30 individual values. Thus for a given
+molecule it is possible to generate approximately 300 descriptor
+values. Every descriptor implements an interface allowing them to
+return values in a uniform manner, viz., via a
+\texttt{DescriptorValue} object. This class can represents the result
+of a descriptor calculation and includes both the descriptor values
+(via a \texttt{IDescriptorResult} object) as well as meta-data on the
+descriptor itself via a \texttt{DescriptorSpecification} object (see
+below for more details). The descriptor classes are designed such that
+in the event of a calculation error (invalid, missing or nonsensical
+input), null values are returned, rather than throwing an
+exception. This is useful during automated descriptor calculation,
+where one desires a ``rectangular'' descriptor matrix, from which one
+could filter out the undefined descriptors. While the user can pick
+and choose specific descriptors to calculate, the API provides a class
+(\texttt{DescriptorEngine}) to specifically evaluate all or a subset
+of descriptors for a molecule at one go.  Currently the descriptor
+implementations are not threadsafe, making them unsuitable for
+parallelization. This is a drawback since many of the descriptors can
+be slow. For example, the $\chi$ descriptors \cite{Kier:1976aa} are
+implemented using SMARTS matching to identify the various paths and
+chains described by Kier et al, rather than explicitly walking the
+molecular graph.
+
 \subsection{Comparing implementations}
@@
 
-One can also directly compare descriptor values generated using
-different implementations.
-
 A number of physicochemical descriptors commonly employed in
@@
 values from the CDK, ChemAxon and ACD Labs for a set of 10,000
-molecules taken from the logPstar dataset [REF].
+molecules taken from the logPstar dataset.
 
@@
 
-\section{Running Notes}
-\begin{itemize}
-\item Survey of descriptor implementations (OSS and non-OSS)
-  \begin{itemize}
-  \item MOE
-  \item OEChem/GraphTK
-  \end{itemize}
-\item Go into more detail of OSS descriptor implementations
-  \begin{itemize}
-  \item CDK (describe architecture), mention wrappers around the
-    library (CDKDescUI and batch mode)
-  \item PaDEL
-  \end{itemize}
-\end{itemize}
 \clearpage
@@
 
+\ctable[caption={A comparison of the VABC descriptor for molecular
+  volume (as implemented by the CDK) and other approximate
+  implementations of molecular volume (as implemented by MOE 2008.10)},cap={},label={fig:molvol},botcap,figure]
+{c}
+{}
+{
+  \includegraphics[width=\linewidth]{vol-moe-cdk}
+}
+
 \ctable[caption={A comparison of Topological Polar Surface Area values

### 2012-04-17 paper.tex
Updated some figure

@@
 \ctable[caption={A comparison of Topological Polar Surface Area values
-generated using the CDK and ACD Labs software, for 57,600 molecules
+generated using the CDK and ACD Labs software, for 57,857 molecules
 taken from Pubchem AID 1996},cap={},label={fig:tpsa},botcap,figure]

### 2012-04-17 code/code.R

@@ -1,0 +1,39 @@
+library(lattice)
+library(reshape)
+
+acd <- read.table('all-clean.acd', sep='\t', header=TRUE, as.is=TRUE)
+cdk <- read.table('all-clean.cdk', sep='\t', header=TRUE, as.is=TRUE)
+
+tmp <- data.frame(ACD=acd$ACD_RuleOf5_PSA, CDK=cdk$TopoPSA)
+xyplot(ACD ~ CDK, tmp, type=c('p', 'g'), col='black', cex=0.25,
+       xlab=list(label='CDK TPSA', cex=1.2),
+       ylab=list(label='ACD TPSA', cex=1.2),
+       panel=function(x,y,...) {
+         panel.xyplot(x,y,...)
+         panel.abline(0, 1, col='red')                  
+       }
+         )
+dev.copy(pdf, '../tpsa-cdk-acd.pdf', 6,6);dev.off()
+dev.copy(png, '../tpsa-cdk-acd.png', 400,400);dev.off()
+
+
+## volumes
+moe <- read.csv('vol.moe', header=TRUE, as.is=TRUE)
+cdk <- read.table('vol.cdk', header=TRUE, as.is=TRUE, sep=' ')
+names(moe) <- c('sid', 'vdw_vol', 'vol')
+names(cdk) <- c('sid', 'VABC')
+tmp <- merge(moe, cdk, by='sid')
+tmp <- tmp[ complete.cases(tmp), ]
+tmp <- melt(tmp, measure.vars=c('vdw_vol', 'vol'), id.vars='VABC')
+xyplot(value ~ VABC | variable, tmp,
+       type=c('p', 'g'), col='black', cex=0.25,
+       xlab=list(label='CDK VABC', cex=1.2),
+       ylab=list(label='MOE', cex=1.2),
+       xlim=range(c(tmp$VABC, tmp$value)),
+       ylim=range(c(tmp$VABC, tmp$value)),       
+        panel=function(x,y,...) {
+         panel.xyplot(x,y,...)
+         panel.abline(0, 1, col='red')                  
+       })
+dev.copy(pdf, '../vol-moe-cdk.pdf', 8,4);dev.off()
+dev.copy(png, '../vol-moe-cdk.png', 600,300);dev.off()

### 2012-04-17 paper.tex
Updated some text

@@
 stability\cite{Gupta:2010uq}. Their results indicated very similar
-performance between the two implementations.
+performance between the two implementations. In general, one would
+expect a high degree of correlation between different implementations
+of well-defined descriptors. Here ``well defined'' indicates that the
+underlying parametrization or chemistry model is completely specified,
+so that different implementations use them identically. For example,
+correct implementations of the TPSA descriptor are expected to
+correlate very well, and this is shown in Figure \ref{fig:tpsa} for
+the ACD Labs and CDK implementations.
 

### 2012-04-17 paper.tex
Added some running notes for a new section
Updated descriptor table

@@
 
+\section{What is a Useful Descriptor?}
+\label{sec:what-usef-descr}
+
+ideal characteristics fo a descriptor
+security
+interpretability
+modeling
+
 \section{Descriptor Implementations}
@@
 \raggedright{Keys from Xue et al\ \cite{Godden:2000qf}} \NN
+MOLCONN-Z & Commercial & No &  Yes & topological &  \NN
 OpenBabel & GPL & Yes & & \raggedright{} & \raggedright{} \NN
+PowerMV \cite{Liu:2005ab} & GPL &  & Yes & \raggedright{constitutional, atom pairs,
+  Burden numbers, pharmacophoric} & \raggedright{} \NN
 RDKit & New BSD & Yes & & \raggedright{Constitutional, topological,

### 2012-04-17 paper.tex
Added a section on basis for a good descriptor

--- 2012-04-17 paper.tex
@@
 predictive modeling applications
-\cite{Garcia-Domenech:2008aa,Randic:2001ac,Randic:2001ad,Besalu:2001aa,jcics:1995:35:272,Kier:1986ae},
+\cite{Garcia-Domenech:2008aa,Randic:2001ad,Besalu:2001aa},
 a key drawback is their lack of interpretability. Given the abstract
@@
 be used in subsequent predictive models.
+
+\section{What is a Useful Descriptor?}
+\label{sec:what-usef-descr}
+
+Clearly, one can calculate a huge variety of descriptors, many of
+which will be correlated with others. It is then critical to ask, what
+makes a descriptor useful? Fundamentally, a descriptor must correlate
+structural features with some physicochemical property and show
+minimal correlation with other descriptors. In addition, a generally
+useful descriptor will be applicable to a broad class of
+molecules. Furthermore, a descriptor should show minimal
+degeneracy. That is, the descriptor should generate different values
+for structurally different molecules, even if the structural
+differences are small. An example is the case of stereochemistry. Most
+2D descriptors that are not designed to take this into account will
+generate the same numerical values for all the stereoisomers of a
+given chiral molecule. We would consider such a descriptor to be
+degenerate. In addition to degeneracy, a descriptor should be
+continuous, in the sense, that small structural changes should lead to
+small changes in the value of the descriptor.
+
+The requirements listed here ensures that subsequent analyses
+involving descriptors are robust. But in addition to numerical
+robustness, it is important for amny scenarios that the descriptor
+have some form of physical interpretability.  For example, a model
+that employs descriptors to predict cytoxicity will be most useful if
+it is numerically accurate as well as providing insight into which
+structural features confer toxicity. In such a scenario, abstract
+descriptors do not allow us to provide a physical explanation of the
+models predictions. Another scenario is the ability to map descriptor
+values back to the structure for visualization purposes such as the
+``Glowing Molecule'' depictions of Segall et al
+\cite{Segall:2009fk}. Such visualizations are practical only when
+descriptor values can be related to structural features. Examples
+include surface based descriptors such as CPSA's as well as explicit
+structural descriptors such as paths, chains etc.
+
+However, there are certain scenarios where some of the above
+characteristics are \emph{undesirable}. One such scenario is in the
+safe exchange of structures, such that two parties can perform
+analyses on the structure-derived data, without having to know the
+structures explicitly. In such cases, the ability to go from a set of
+descriptor values back to a chemical structure should be minimized. It
+has not been conclusively proven that such a ``one-way'' descriptor
+can be defined. For example Masek et al \cite{Masek:2008aa} have shown
+that one can regenerate a target structure (or substructural features)
+or a set of closely related analogs given a combination of descriptors
+such as BCUTs. Tetko et al \cite{Tetko:2005fk} suggest an approach to
+prevent reverse engineering bu avoid the use of descriptors and
+instead sharing structures of related molecules, rather than the
+molecules of interest. One could also employ degenerate descriptors,
+which would represent a large family of molecules, such that the
+target molecule would be hidden. This is likely unreliable as it would
+be difficult to perform analyses (e.g., predtive models) with such
+descriptors.
 
@@
 \raggedright{Keys from Xue et al\ \cite{Godden:2000qf}} \NN
+Molconn-Z & Commercial & No &  Yes & topological &  \NN
 OpenBabel & GPL & Yes & & \raggedright{} & \raggedright{} \NN
+PowerMV \cite{Liu:2005ab} & GPL &  & Yes & \raggedright{constitutional, atom pairs,
+  Burden numbers, pharmacophoric} & \raggedright{} \NN
 RDKit & New BSD & Yes & & \raggedright{Constitutional, topological,

### 2012-04-18 paper.tex
Added summary section

@@
 and docking) to more abstract numerical descriptions derived from 2D
-or 3D representations which can be useful is statistical approaches.
-
-It is now possible to evaluate thousands of numerical descriptors of
-chemical structure. As will be discussed later, many of these
-descriptors are closely related - one can be substituted for
+or 3D representations which can be useful is statistical
+approaches. It is now possible to evaluate thousands of numerical
+descriptors of chemical structure. As will be discussed later, many of
+these descriptors are closely related - one can be substituted for
 another. The choice of descriptors is a well known problem and given a
@@
 
-Clearly, one can calculate a huge variety of descriptors, many of
-which will be correlated with others. It is then critical to ask, what
+Given that we can generate thousands of descriptors,  it is critical to ask, what
 makes a descriptor useful? Fundamentally, a descriptor must correlate
@@
 class of applications for descriptor generation and allow one to
-easily generate descriptors from multiple sources.
-
-Given the focus of this article on Open Source implementations, we
-have noted the availability in Table \ref{tab:impl}. While a number of
+easily generate descriptors from multiple sources. Given the focus of
+this article on Open Source implementations, we have noted the
+availability in Table \ref{tab:impl}. While a number of
 implementations are not strictly Open Source according to OSI
@@
 decision tree models to predict human liver microsomal metabolic
-stability\cite{Gupta:2010uq}. Their results indicated very similar
+stability \cite{Gupta:2010uq}. Their results indicated very similar
 performance between the two implementations. In general, one would
@@
 between themselves. Figure \ref{fig:logp} compares computed log P
-values from the CDK, ChemAxon and ACD Labs for a set of 10,000
-molecules taken from the logPstar dataset.
+values from the CDK (specifically, an implementation of the XlogP
+method), ChemAxon and ACD Labs for a set of 10,000 molecules taken
+from the logPstar dataset.
 
@@
 \label{sec:summary}
+
+Molecular descriptors play a fundamental role in cheminformatic and
+chemometric analyses and as we have described in this article it is
+possible to evaluate thousands of descriptors using a variety of
+software. Though the bulk of descriptors are well defined in the
+literature, multiple implementations of the same descriptor can yield
+different results. These differences arise from differences in the
+underlying chemistry models and reference data used by the
+implementations. As shown by Gupta et al \cite{Gupta:2010uq},
+the performance of predictive models developed using commercial or
+Open Source descriptor implementations is very similar. While
+different tools differ in the specific number and types of descriptors
+that they calculate, the fact many descriptors are correlated with
+others suggests that it does not matter too much which set of
+descriptors are used in an application. Of course, this cannot be a
+general rule - specific applications may require a specific set of
+descriptors. This is not to say that all descriptor tools exhibit
+similr performance. For example, it is clear from Figure \ref{fig:logp}
+that the ACD Labs implementation of log P performs significantly
+better than other implementations. Though this is not a completely
+rigorous comparison (the methods underlying the ACD Labs, CDK and
+ChemAxon implementations are different), it does highlight that
+certain implementations of a descriptor can fare better than others -
+especially those cases where the descriptor is based on a predictive
+model.
+
+Given these challenges it is imperative that descriptor tools provide
+access to version information. This allows the user to provide an
+exact specification of how the descriptor was calculated. However, one
+aspect that has not been standardized across different implementations
+is the naming scheme. In other words, tools that evaluate the same
+descriptor may name them in a similar but not exactly identical
+manner. This makes automated comparisons and merging of descriptors
+from different sources problematic. To address the CDK has developed
+the concept of a ``descriptor specification'' which is associated with
+a descriptor value and includes information on the vendor, descriptor
+title and a reference to an entry in a descriptor dictionary that
+contains more details of the descriptor in RDF. While the extra
+descriptor metadata is currently used to implement a simple
+classification scheme, it is conceivable that in the future the
+specification approach could be adopted by multiple vendors, allowing
+for automated reasoning over descriptor implementations.
+
+In summary, there is no dearth of tools to generate molecular
+descriptors. Many of them are available under liberal Open Source
+licenses, though these implementations do not necessrily cover all the
+descriptors described to date. The tools range from toolkits (which
+implies that one must write a program to generate descriptors) to
+self-contaiend GUI or command line tools. Given the caveats regarding
+implementation specific differences, it is important to keep track of
+provenance information when calculating descriptors to ensure
+reproducibility.
 

### 2012-04-23 paper.tex
Fixed a few spelling errors

@@
   descriptors and then describe applications and toolkits that can be
-  employed to evaluate them. We highight a number of issues
+  employed to evaluate them. We highlight a number of issues
   surrounding molecular descriptor calculations such as versioning and
@@
 numerical forms range from a set of 3D coordinates (which coupled with
-appropiate atom types, is sufficient for methods such QM approaches
+appropriate atom types, is sufficient for methods such QM approaches
 and docking) to more abstract numerical descriptions derived from 2D
@@
 a challenge for descriptor development. There have been reports that
-employed tradtional small-molecule descriptors
+employed traditional small-molecule descriptors
 \cite{Mattioni:2002dq,Katritzky:1998cr} that consider the polymer in
@@
 inorganic materials are very similar to constitutional descriptors -
-essentially counts of various crystal parameters such as acccessible
+essentially counts of various crystal parameters such as accessible
 surface areas and volumes
@@
 Sedykh et al \cite{Sedykh:2011fk} who hypothesized that dose response
-data points from high throughput dose-reponse assays could be employed
+data points from high throughput dose-response assays could be employed
 as biological descriptors. They observed that when such descriptors
@@
 involving descriptors are robust. But in addition to numerical
-robustness, it is important for amny scenarios that the descriptor
+robustness, it is important for many scenarios that the descriptor
 have some form of physical interpretability.  For example, a model
@@
 or a set of closely related analogs given a combination of descriptors
-such as BCUTs. Tetko et al \cite{Tetko:2005fk} suggest an approach to
+such as BCUTs. Tetko \textit{et al.} \cite{Tetko:2005fk} suggest an approach to
 prevent reverse engineering bu avoid the use of descriptors and
@@
 implemented using SMARTS matching to identify the various paths and
-chains described by Kier et al, rather than explicitly walking the
+chains described by Kier \textit{et al.}, rather than explicitly walking the
 molecular graph.
@@
 QSAR modeling community lacks such a scheme, though the Chemistry
-Developmen Kit (CDK), a Java library for cheminformatics
+Development Kit (CDK), a Java library for cheminformatics
 \cite{Steinbeck:2006aa} does implement such a scheme. The ``descriptor
@@
 information of the implementation title, identifier and vendor. The
-identifier may or may not incude a version number. The actual details
+identifier may or may not include a version number. The actual details
 of the a descriptor are listed in a dictionary, along with an optional
@@
 descriptors. This is not to say that all descriptor tools exhibit
-similr performance. For example, it is clear from Figure \ref{fig:logp}
+similar performance. For example, it is clear from Figure \ref{fig:logp}
 that the ACD Labs implementation of log P performs significantly
@@
 descriptors. Many of them are available under liberal Open Source
-licenses, though these implementations do not necessrily cover all the
+licenses, though these implementations do not necessarily cover all the
 descriptors described to date. The tools range from toolkits (which
 implies that one must write a program to generate descriptors) to
-self-contaiend GUI or command line tools. Given the caveats regarding
+self-contained GUI or command line tools. Given the caveats regarding
 implementation specific differences, it is important to keep track of

### 2012-04-23 paper.tex
Added the UML diagram

@@
 values. Every descriptor implements an interface allowing them to
-return values in a uniform manner, viz., via a
-\texttt{DescriptorValue} object. This class can represents the result
+return values in a uniform manner, viz., via a \texttt{DescriptorValue}
+object (see Figure~\ref{fig:cdkapi}). This class can represents the result
 of a descriptor calculation and includes both the descriptor values
@@
 
+\ctable[caption={UML diagram of the CDK descriptor API in version 1.4.},cap={},label={fig:cdkapi},botcap,figure]
+{c}
+{}
+{
+  \includegraphics[width=\linewidth]{iDescriptor.pdf}
+}
+
 \ctable[caption={A comparison of Topological Polar Surface Area values

### 2012-05-14 paper.tex
Tuned the introduction and description of the descriptor types a bit:
- minor textual editing
- added subsections for easier reading
- added a section on crystal structure descriptors
- clarified that VABC is not a geometrical but a topological descriptor,
  following the given definitions of those

@@
 other words, explicit representations of chemical structure such as
-SMILES, SD files and so on. While these descriptions are vital to
+SMILES, MDL/Symyx SD files and so on. While these descriptions are vital to
 modern chemical information systems, they do not necessarily allow
@@
 numerical forms range from a set of 3D coordinates (which coupled with
-appropriate atom types, is sufficient for methods such QM approaches
+appropriate atom types, is sufficient for methods such quantum mechenical (QM) approaches
 and docking) to more abstract numerical descriptions derived from 2D
@@
 descriptors of chemical structure. As will be discussed later, many of
-these descriptors are closely related - one can be substituted for
-another. The choice of descriptors is a well known problem and given a
+these descriptors are closely related or capture the same information,
+causing that one can be substituted for another. The choice of descriptors is a well-known problem and given a
 large collection of them, approaches to identify a suitable subset
-have been discussed extensively in the literature
-\cite{Miller:2002aa,Kohavi:1997gf}.
+have been discussed extensively in the literature~\cite{Miller:2002aa,Kohavi:1997gf}.
 
@@
 libraries (which require one to write a program to use them) or
-complete applications (GUI or command line). As a result, not only
+complete applications (graphical user interface (GUI) or command line). As a result, not only
 must one choose one or more descriptors that are relevant to the
@@
 understand \emph{why} two implementations of the same descriptor can
-lead to different results - the primary reason being the chemistry
+lead to different results - the primary reasons being that the chemistry
 model that underlies the framework or toolkit used to implement the
-descriptor. For example, a descriptor that calculates the number of
+descriptor is different, and that used data such as isotopic masses
+are different.
+
+For example, a descriptor that calculates the number of
 aromatic atoms may be implemented using two toolkits with differing
@@
 concepts (e.g., atomic weights), slight differences in these types of
-input data can lead to differences in the final descriptor value. As a
+input data can lead to differences in the final descriptor value~\cite{Guha:2006ac}. As a
 result, in most cases, two implementations of a descriptor do not
@@
 possible (it is usually more difficult in the cases of commercial
-implementations for which source code is not available).
+implementations for which source code is not available)~\cite{Spjuth2010}.
 
@@
 We have noted that a molecular structure can be characterized using a
-numebr of numerical descriptors. In this section we describe a
+number of numerical descriptors. In this section we describe a
 categorization of descriptors. Admittedly, the grouping is somewhat
@@
 for polymers or protein structures.
+Broadly, one can classify descriptors based on the nature of
+structural information they require.
 
-Broadly, one can classify descriptors based on the nature of
-structural information they require. Constitutional descriptors only
+Constitutional descriptors only
 require atom and bond labels and usually represent counts of different
@@
 precalculated surface area values derived from a list of functional
-groups.  Molecular volume is also a useful physical descriptor and
-both analytical \cite{Eisenhaber:1995qf} and approximate methods are
-available to evaluate it. A very fast method is the VABC approach
-\cite{Zhao:2003fk} that is based on group contributions. It has been
+groups.
+
+It is important to note that there is a difference between the information
+a descriptor captures and how it is calculated. The above categories are
+based on input requirements. This gives rise to situations where
+geometrical molecular volume property can be described by geometrical descriptors,
+e.g. using an analytical approaches ~\cite{Eisenhaber:1995qf}, and
+with approximate, constitutional descriptors, such as the very fast algorithm
+use by the VABC approach~\cite{Zhao:2003fk} which is based on group contributions.
+The latter descriptor has been
 implemented in the CDK and correlates well with other approximations
@@
 
-Molecular shape is a key factor in ligand-receptor interactions and is
+Another important class of geometrical descriptors are those that
+describe molecular shape. This property is a key factor in ligand-receptor interactions and is
 an important approach to virtual screening. A variety of descriptors
@@
 
-Finally, quantum mechanical calculations can also generated
-descriptors that can be applied to problems such as QSAR models
+The last category consists of QM descriptors. These
+descriptors can also be applied to problems such as QSAR models
 (though of course the time complexity of QM derived descriptors do not
@@
 
-\textbf{Fingerprints} Traditionally, fingerprint descriptors are
-represented in the form of bit strings. Binary fingerprints can be
+\subsection{Fingerprints}
+
+Besides the numerical descriptors outlined so far, another useful
+approach are those of fingerprints. Traditionally, fingerprint descriptors are
+represented in the form of bit strings. These binary fingerprints can be
 divided into hashed fingerprints, where substructures (such as paths
@@
 
+\subsection{Describing other chemistry}
+
 We next consider descriptors that are designed for specific classes of
@@
 descriptors characterize the periodic voids of porous materials via
-tesselation methods \cite{Anurova:2009ly,Carr:2009kx}
+tesselation methods \cite{Anurova:2009ly,Carr:2009kx}.
+
+\textbf{Crystal structures} Other methods developed for crystal structures
+have been developed too, for example, using a radial distribution
+functions~\cite{Willighagen2005}, which capture the packing of organic
+crystal structures. Such approaches have been used to cluster
+crystal structures~\cite{Wehrens2006} and compare properties~\cite{Willighagen2007}.
+% FIXME: add recent work
 
@@
 differences are small. An example is the case of stereochemistry. Most
-2D descriptors that are not designed to take this into account will
+constitutional and topological descriptors that are not designed to take this into account will
 generate the same numerical values for all the stereoisomers of a
@@
 such as BCUTs. Tetko \textit{et al.} \cite{Tetko:2005fk} suggest an approach to
-prevent reverse engineering bu avoid the use of descriptors and
+prevent reverse engineering but avoid the use of descriptors and
 instead sharing structures of related molecules, rather than the

### 2012-05-14 paper.tex
More bits

@@
 
+\textbf{Atom and bond descriptors} It should also be noted that various
+toolkits also calculate descriptors for chemical entities like atoms and
+bonds. These descriptors include simple element property descriptors,
+but also geometrical descriptor, such as radial distribution functions
+that describe the chemical environment of those atoms. These,
+for example, have been used to predict proton shift in 
+nuclear magnetic resonance~\cite{AiresDeSousa2002}. Similarly, bond
+descriptors can be used to predict bond properties.
+
 \section{What is a Useful Descriptor?}
@@
 cheminformatics toolkits providing implementations of a common set of
-descriptors and fingerprints. Table \ref{tab:impl} summarizes the
+descriptors and fingerprints. Table~\ref{tab:impl} summarizes the
 descriptor classes implemented by several toolkits and
-applications. Note that that due to space limitations this is not a
+applications. Note that due to space limitations this is not a
 comprehensive list of implemented descriptors across different
@@
 (\url{http://rguha.net/code/java/cdkdesc.html}) is a Java Swing
-application that exposes the CDK descriptor and fingerprint API's. It
+application that exposes the Chemistry Development Kit (CDK) descriptor and fingerprint API's. It
 allows the users to specify a SMILES or SD file and evaluate a
@@
 class of applications for descriptor generation and allow one to
-easily generate descriptors from multiple sources. Given the focus of
+easily generate descriptors from multiple sources~\cite{Truszkowski2011}. Given the focus of
 this article on Open Source implementations, we have noted the
@@
 and whole molecule descriptors. There are a total of 30, 6 and 51
-descriptors for atoms, bonds and molecules respectively. Note that
-each individual descriptor may calculate multiple values. For example,
-the CPSA descriptor evaluates 30 individual values. Thus for a given
+descriptor algorithms for atoms, bonds and molecules respectively. Note that
+each individual descriptor algorithm may calculate multiple descriptors. For example,
+the CPSA descriptor evaluates 30 individual descriptor values. Thus for a given
 molecule it is possible to generate approximately 300 descriptor
-values. Every descriptor implements an interface allowing them to
+values.
+
+Every descriptor implements an interface allowing them to
 return values in a uniform manner, viz., via a \texttt{DescriptorValue}
@@
 same descriptor can generate different values. Thus to maintain
-reproducibility in sudies that employ molecular descriptors, some form
+reproducibility in studies that employ molecular descriptors, some form
 of versioning is crucial. The simplest form of version is to manually
@@
 refer to descriptors via a standardized naming scheme. Currently the
-QSAR modeling community lacks such a scheme, though the Chemistry
-Development Kit (CDK), a Java library for cheminformatics
+QSAR modeling community lacks such a scheme, though the CDK, a Java library for cheminformatics
 \cite{Steinbeck:2006aa} does implement such a scheme. The ``descriptor
@@
 enable easier interoperability between descriptor implementations and
-enhance reproducibility.
+enhance reproducibility. The framework is, however, actively used by
+software using the CDK descriptors, such as
+Bioclipse~\cite{Spjuth2010} and Ambit~\cite{Jelizakova2011}. Moreover,
+the ideas have been adopted by the CHEMINF ontology for
+cheminformatics~\cite{Hastings2011}, which is used by the CHESS
+framework which also uses the CDK for descriptor calculations~\cite{Chepelev2011b}.
 

### 2012-05-18 paper.tex
Added the new screenshots to the chapter

@@
 applications that allow easy descriptor calculations. Examples include
-Bioclipse \cite{Spjuth:2007aa}, MOE (Chemical Computing Group) and
+Bioclipse \cite{Spjuth:2007aa} (see Figure~\ref{fig:tools}), MOE (Chemical Computing Group) and
 Maestro (Schr\"{o}dinger, Inc.). Workflow tools are also a useful
 class of applications for descriptor generation and allow one to
-easily generate descriptors from multiple sources~\cite{Truszkowski2011}. Given the focus of
+easily generate descriptors from multiple sources~\cite{Truszkowski2011} (see Figure~\ref{fig:tools}). Given the focus of
 this article on Open Source implementations, we have noted the
@@
 same descriptor can generate different values. Thus to maintain
-reproducibility in studies that employ molecular descriptors, some form
+reproducibility in studies that employ moleculad2a28dc0c273a4e060c2fea503adc36er descriptors, some form
 of versioning is crucial. The simplest form of version is to manually
@@
 
+\ctable[caption={Four tools to calculate descriptors: Bioclipse (top left), which allows selecting descriptor
+implementations from various independent tools, CDK-Taverna (top right), which is an extension to Taverna
+to calculate descriptors with the CDK, Ambit2 (bottom left), which implements the OpenTox API providing
+a REST-based API for descriptor calculation and wraps various descriptor calculation tools,
+and ....},cap={},label={fig:tools},botcap,figure]
+{c}
+{}
+{
+  \includegraphics[width=0.48\linewidth]{bioclipse}
+  \includegraphics[width=0.48\linewidth]{cdktaverna} \\
+  \includegraphics[width=0.48\linewidth]{ambitOnline}
+}
+
 \ctable[caption={UML diagram of the CDK descriptor API in version 1.4.},cap={},label={fig:cdkapi},botcap,figure]

### 2012-05-18 paper.tex
Added a bit of text on online APIs

@@
 
+There are also web interfaces wrapping descriptor calculation tools
+and some of these interfaces have a programmable feature. For example,
+AMBIT2 implement the OpenTox API providing a REST-like interface
+to descriptor calculation~\cite{Jeliazkova2011,Hardy2010} (see Figure~\ref{fig:tools}). Another
+online application programming interface for descriptor calculation
+is provided by the SADI platform, which, like AMBIT2, uses semantic
+web technologies for descriptor calculation~\cite{Chepelev2011b}.
+
 In addition to libraries and their wrappers there are various GUI

### 2012-07-03 paper.tex
some minor modifications, ready for submission

@@
   Numerical characterization of molecular structure is a first step in
-  every computational analysis of molecular data. These numerical
+  many computational analysis of chemical structure data. These numerical
   representations, termed descriptors, come in many forms, ranging
@@
 Computational methods play an important role in many chemical
-disciplines ranging from drug discovery to materials science. While
-there are a variety of techniques that differ in terms of
+disciplines ranging from drug discovery to materials science.
+There are a plethora of techniques that differ in terms of
 computational complexity, time requirements and so on. However the
 common requirement underlying all these methods is a formal
-description of a the molecular structure. Of course, there are many
+description of a the molecular structure. There are many
 ways to ``describe'' a molecule. A common approach is to describe the
@@
 numerical forms range from a set of 3D coordinates (which coupled with
-appropriate atom types, is sufficient for methods such quantum mechenical (QM) approaches
-and docking) to more abstract numerical descriptions derived from 2D
-or 3D representations which can be useful is statistical
-approaches. It is now possible to evaluate thousands of numerical
-descriptors of chemical structure. As will be discussed later, many of
-these descriptors are closely related or capture the same information,
-causing that one can be substituted for another. The choice of descriptors is a well-known problem and given a
-large collection of them, approaches to identify a suitable subset
-have been discussed extensively in the literature~\cite{Miller:2002aa,Kohavi:1997gf}.
+appropriate atom types, is sufficient for methods such as quantum
+mechanical (QM) approaches and docking to more abstract numerical
+descriptions derived from 2D or 3D representations which can be useful
+in statistical approaches. It is now possible to evaluate thousands of
+numerical descriptors of chemical structure. As will be discussed
+later, many of these descriptors are closely related or capture the
+same information, allowing one to be substituted for another. The
+choice of descriptors is a well-known problem and given a large
+collection of them, approaches to identify a suitable subset have been
+discussed extensively in the
+literature~\cite{Miller:2002aa,Kohavi:1997gf}.
 
@@
 libraries (which require one to write a program to use them) or
-complete applications (graphical user interface (GUI) or command line). As a result, not only
-must one choose one or more descriptors that are relevant to the
-problem at hand, but one must be concerned about how they
-are calculated and whether such a calculation can be reproduced across
-different implementations of those descriptors. It is easy to
-understand \emph{why} two implementations of the same descriptor can
-lead to different results - the primary reasons being that the chemistry
-model that underlies the framework or toolkit used to implement the
-descriptor is different, and that used data such as isotopic masses
-are different.
-
-For example, a descriptor that calculates the number of
-aromatic atoms may be implemented using two toolkits with differing
-aromaticity models, and hence it is possible that the values generated
-by the two implementations will differ. Other sources of differences
-include parameters that may be involved in the descriptor calculation
-and reference data values (such as atomic radii, electronegativity
-values) that are employed during descriptor calculation. While most
-implementations will employ the same data sources for standard
-concepts (e.g., atomic weights), slight differences in these types of
-input data can lead to differences in the final descriptor value~\cite{Guha:2006ac}. As a
-result, in most cases, two implementations of a descriptor do not
-usually give the exact same value, though they are usually quite
-similar. Explicitly explaining the differences may or may not be
-possible (it is usually more difficult in the cases of commercial
-implementations for which source code is not available)~\cite{Spjuth2010}.
+complete applications (graphical user interface (GUI) or command
+line). As a result, not only must one choose one or more descriptors
+that are relevant to the problem at hand, but one must be concerned
+about how they are calculated and whether such a calculation can be
+reproduced across different implementations of those descriptors. It
+is easy to understand \emph{why} two implementations of the same
+descriptor can lead to different results - the primary reasons being
+that the chemistry model that underlies the framework or toolkit used
+to implement the descriptor is different. For example, a descriptor
+that calculates the number of aromatic atoms may be implemented using
+two toolkits with differing aromaticity models, and hence it is
+possible that the values generated by the two implementations will
+differ. Other sources of differences include parameters that may be
+involved in the descriptor calculation and reference data values (such
+as atomic radii, electronegativity values) that are employed during
+descriptor calculation. While most implementations will employ the
+same data sources for standard concepts (e.g., atomic weights), slight
+differences in these types of input data can lead to differences in
+the final descriptor value~\cite{Guha:2006ac}. As a result, in most
+cases, two implementations of a descriptor do not usually give the
+exact same value, though they are usually quite similar. Explicitly
+explaining the differences may or may not be possible (it is usually
+more difficult in the cases of commercial implementations for which
+source code is not available)~\cite{Spjuth2010}.
 
@@
 
-We have noted that a molecular structure can be characterized using a
-number of numerical descriptors. In this section we describe a
+We have noted that a molecular structure can be characterized using
+many numerical descriptors. In this section we describe a
 categorization of descriptors. Admittedly, the grouping is somewhat
 arbitrary but serves as a useful guide. In addition, the
-categorization also considers the nature of the chemical structure
-being considered - some descriptors are only useful when applied to
-small molecules, whereas other descriptors are defined specifically
-for polymers or protein structures.
-Broadly, one can classify descriptors based on the nature of
-structural information they require.
+categorization also takes into account the nature of the chemical
+structure being considered - some descriptors are only useful when
+applied to small molecules, whereas other descriptors are defined
+specifically for polymers or protein structures.  Broadly, one can
+classify descriptors based on the nature of structural information
+they require.
 
-Constitutional descriptors only
-require atom and bond labels and usually represent counts of different
-types of atoms or bonds. While very simplistic they can play a useful
-role in a variety of applications ranging from summaries of
-physicochemical properties to predictive modeling
-\cite{Bender:2005aa}.
+Constitutional descriptors only require atom and bond labels and
+usually represent counts of different types of atoms or bonds. While
+very simplistic they can play a useful role in a variety of
+applications ranging from summaries of physicochemical properties to
+predictive modeling \cite{Bender:2005aa}.
 
 Topological descriptors take into account connectivity along with atom
-and bond labels. These descriptors essentially consider the molecule
-as a labeled graph and characterize it using graph invariants. In
-addition to graph invariants, topological descriptors based on
-information theory such as entropy \cite{Dehmer:2009uq}. An advantage
-of these descriptors is that do not require any intensive
-pre-processing steps such as 3D coordinate generation and
-conformational analysis. In addition, many graph invariants can be
-rapidly computed for graphs corresponding to small molecule structures
-(even though they may be computationally intractable on much larger
-graphs). While this class of descriptors have been used extensively in
-predictive modeling applications
-\cite{Garcia-Domenech:2008aa,Randic:2001ad,Besalu:2001aa},
-a key drawback is their lack of interpretability. Given the abstract
-nature of many graph invariants and information theoretic descriptors,
-their use in predictive models makes it difficult to explain the model
-in simple physicochemical terms, though various attempts have been  
-made at interpretations \cite{Todeschini:1975dq,Stanton:2003aa}.
+and bond labels. They consider the molecule as a labeled graph and
+characterize it using graph invariants. In addition to graph
+invariants, topological descriptors based on information theory such
+as entropy have also been defined\cite{Dehmer:2009uq}. An advantage of
+these descriptors is that do not require any intensive pre-processing
+steps such as 3D coordinate generation and conformational analysis. In
+addition, many graph invariants can be rapidly computed for graphs
+corresponding to small molecule structures (even though they may be
+computationally intractable on much larger graphs). While this class
+of descriptors have been used extensively in predictive modeling
+applications
+\cite{Garcia-Domenech:2008aa,Randic:2001ad,Besalu:2001aa}, a key
+drawback is their lack of interpretability. The abstract nature of
+many graph invariants and information theoretic descriptors, their use
+in predictive models makes it difficult to explain the model in simple
+physicochemical terms, though various attempts have been made at
+interpretations \cite{Todeschini:1975dq,Stanton:2003aa}.
 
 Geometric descriptors require a 3D conformation as input and therefore
-are not as fast as topological descriptors which do not require the 3D
-coordinate generation step. Nearly all geometric descriptors work on a
-single conformation and for cases where multiple conformation must be
-considered, an averaging procedure is usually employed. Examples of
-these descriptors include the gravitational index
-\cite{Katritzky:1996ly} and moment of inertia descriptors. However,
-geometric descriptors encompass many aspects of a molecular
-structure. For example, a number of geometric descriptors characterize
-the molecular surface. The simplest such example is the surface area
-descriptor (and the corresponding volume descriptor). A simple
-extension of such a descriptor is to characterize distributions of a
-physicochemical property over the molecular surface such as the
-Charged Partial Surface Area descriptors \cite{Stanton:1990aa} that
-characterize the partial charge distribution over the molecular
-surfce. Note that surface descriptor calculations based on analytical
+the computations are not as fast as topological descriptors which do
+not require the 3D coordinate generation step. Nearly all geometric
+descriptors work on a single conformation and for cases where multiple
+conformation must be considered, an averaging procedure is usually
+employed. Examples of these descriptors include the gravitational
+index \cite{Katritzky:1996ly} and moment of inertia descriptors.
+Geometric descriptors encompass many aspects of a molecular
+structure. For example, a number of them characterize the molecular
+surface. The simplest such example is the surface area descriptor (and
+the corresponding volume descriptor). A simple extension of such a
+descriptor is to characterize distributions of a physicochemical
+property over the molecular surface such as the Charged Partial
+Surface Area descriptors \cite{Stanton:1990aa} that characterize the
+partial charge distribution over the molecular surfce. Note that
+surface descriptor calculations based on analytical
 \cite{Connolly:1983aa} or tesselation algorithms
@@
 
-It is important to note that there is a difference between the information
-a descriptor captures and how it is calculated. The above categories are
-based on input requirements. This gives rise to situations where
-geometrical molecular volume property can be described by geometrical descriptors,
-e.g. using an analytical approaches ~\cite{Eisenhaber:1995qf}, and
-with approximate, constitutional descriptors, such as the very fast algorithm
-use by the VABC approach~\cite{Zhao:2003fk} which is based on group contributions.
-The latter descriptor has been
-implemented in the CDK and correlates well with other approximations
-to the molecular volume (Figure \ref{fig:molvol}).
+It is important to note that there is a difference between the
+information a descriptor captures and how it is calculated. The above
+categories are based on input requirements. This gives rise to
+situations where the molecular volume property can be
+described by geometrical descriptors, e.g. using an analytical
+approaches ~\cite{Eisenhaber:1995qf}, and with approximate,
+constitutional descriptors, such as the very fast algorithm use by the
+VABC approach~\cite{Zhao:2003fk} which is based on group
+contributions.  The latter descriptor has been implemented in the CDK
+and correlates well with other approximations to the molecular volume
+(Figure \ref{fig:molvol}).
 
@@
 
-\subsection{Describing other chemistry}
+\subsection{Describing other chemistries}
 
@@
 
-\textbf{Descriptors for polymeric systems} Polymeric systems represent
-a challenge for descriptor development. There have been reports that
+\textbf{Descriptors for polymeric systems} These systems represent a
+challenge for descriptor development. There have been reports that
 employed traditional small-molecule descriptors
@@
 
-\textbf{Crystal structures} Other methods developed for crystal structures
-have been developed too, for example, using a radial distribution
+Other methods developed for crystal structures have been developed
+too, for example, using a radial distribution
 functions~\cite{Willighagen2005}, which capture the packing of organic
-crystal structures. Such approaches have been used to cluster
-crystal structures~\cite{Wehrens2006} and compare properties~\cite{Willighagen2007}.
-% FIXME: add recent work
+crystal structures. Such approaches have been used to cluster crystal
+structures~\cite{Wehrens2006} and compare
+properties~\cite{Willighagen2007}.
 
@@
 target molecule would be hidden. This is likely unreliable as it would
-be difficult to perform analyses (e.g., predtive models) with such
+be difficult to perform analyses (e.g., predictive models) with such
 descriptors.
@@
 same descriptor can generate different values. Thus to maintain
-reproducibility in studies that employ moleculad2a28dc0c273a4e060c2fea503adc36er descriptors, some form
-of versioning is crucial. The simplest form of version is to manually
-note the version of the implementation used for the calculation and
-report it along with the results. Though useful, it is not amenable to
-automation. In some cases, toolkits will provide an API call to
-retrieve the version information, which can then be associated with a
-descriptor calculation.
+reproducibility in studies that employ molecular descriptors, some
+form of versioning is crucial. The simplest form of version is to
+manually note the version of the implementation used for the
+calculation and report it along with the results. Though useful, it is
+not amenable to automation. In some cases, toolkits will provide an
+API call to retrieve the version information, which can then be
+associated with a descriptor calculation.
 

### 2012-07-04 paper.tex
OK, Fig 2 only has three tools

@@
   volume (as implemented by the CDK) and other approximate
-  implementations of molecular volume (as implemented by MOE 2008.10)},cap={},label={fig:molvol},botcap,figure]
+  implementations of molecular volume (as implemented by MOE 2008.10).},cap={},label={fig:molvol},botcap,figure]
 {c}
@@
 
-\ctable[caption={Four tools to calculate descriptors: Bioclipse (top left), which allows selecting descriptor
+\ctable[caption={Three tools to calculate descriptors: Bioclipse (top left), which allows selecting descriptor
 implementations from various independent tools, CDK-Taverna (top right), which is an extension to Taverna
-to calculate descriptors with the CDK, Ambit2 (bottom left), which implements the OpenTox API providing
-a REST-based API for descriptor calculation and wraps various descriptor calculation tools,
-and ....},cap={},label={fig:tools},botcap,figure]
+to calculate descriptors with the CDK, Ambit2 (bottom), which implements the OpenTox API providing
+a REST-based API for descriptor calculation and wraps various descriptor calculation
+tools.},cap={},label={fig:tools},botcap,figure]
 {c}
@@
 generated using the CDK and ACD Labs software, for 57,857 molecules
-taken from Pubchem AID 1996},cap={},label={fig:tpsa},botcap,figure]
+taken from Pubchem AID 1996.},cap={},label={fig:tpsa},botcap,figure]
 {c}

### 2012-07-04 paper.tex
A few spelling fixes

@@
 Surface Area descriptors \cite{Stanton:1990aa} that characterize the
-partial charge distribution over the molecular surfce. Note that
+partial charge distribution over the molecular surface. Note that
 surface descriptor calculations based on analytical
-\cite{Connolly:1983aa} or tesselation algorithms
-\cite{Eisenhaber:1995qf} can be slow and that parametrized methods
+\cite{Connolly:1983aa} or tessellation algorithms
+\cite{Eisenhaber:1995qf} can be slow and that parameterized methods
 such as the Topological Polar Surface Area (TPSA) \cite{Ertl:2000aa}
@@
 representations is the use of $\alpha$ shapes
-\cite{Edelsbrunner:1994aa}, which represent a parametrization of the
+\cite{Edelsbrunner:1994aa}, which represent a parameterization of the
 convex hull derived from the molecular coordinates. This approach has
@@
 descriptors characterize the periodic voids of porous materials via
-tesselation methods \cite{Anurova:2009ly,Carr:2009kx}.
+tessellation methods \cite{Anurova:2009ly,Carr:2009kx}.
 
@@
 have some form of physical interpretability.  For example, a model
-that employs descriptors to predict cytoxicity will be most useful if
+that employs descriptors to predict cytotoxicity will be most useful if
 it is numerically accurate as well as providing insight into which
@@
 but unlikely, that implementations of the same descriptor using
-different cheminformatcs toolkits will be identical. Gupta et al
+different cheminformatics toolkits will be identical. Gupta et al
 employed SMARTS based descriptors from the CDK and MOE to develop
@@
 generated using the CDK and ACD Labs software, for 57,857 molecules
-taken from Pubchem AID 1996.},cap={},label={fig:tpsa},botcap,figure]
+taken from PubChem AID 1996.},cap={},label={fig:tpsa},botcap,figure]
 {c}

### 2012-07-05 paper.tex
Some minor updates. Added Bocatelli reference

@@
 generated using the CDK, ACD Labs and ChemAxon software, for 10,000 molecules
-taken from the proprietary logPstar dataset [REF?]},cap={},label={fig:logp},botcap,figure]
+taken from the proprietary logPstar dataset.},cap={},label={fig:logp},botcap,figure]
 {c}

### 2012-07-05 paper.tex
Added LICSS

@@
 Bioclipse \cite{Spjuth:2007aa} (see Figure~\ref{fig:tools}), MOE (Chemical Computing Group) and
-Maestro (Schr\"{o}dinger, Inc.). Workflow tools are also a useful
+Maestro (Schr\"{o}dinger, Inc.) as standalone tools, but also
+as a plugin to Microsoft Excel, called LICSS~\cite{Lawson2011}.
+Workflow tools are also a useful
 class of applications for descriptor generation and allow one to
-easily generate descriptors from multiple sources~\cite{Truszkowski2011} (see Figure~\ref{fig:tools}). Given the focus of
+easily generate descriptors from multiple sources~\cite{Truszkowski2011}
+(see Figure~\ref{fig:tools}). Given the focus of
 this article on Open Source implementations, we have noted the
@@
 
-\ctable[caption={Three tools to calculate descriptors: Bioclipse (top left), which allows selecting descriptor
+\ctable[caption={Four tools to calculate descriptors: Bioclipse (top left), which allows selecting descriptor
 implementations from various independent tools, CDK-Taverna (top right), which is an extension to Taverna
-to calculate descriptors with the CDK, Ambit2 (bottom), which implements the OpenTox API providing
+to calculate descriptors with the CDK, Ambit2 (bottom left), which implements the OpenTox API providing
 a REST-based API for descriptor calculation and wraps various descriptor calculation
-tools.},cap={},label={fig:tools},botcap,figure]
+tools. The bottom right screenshot shows the Microsoft Excel plugin LICSS
+that uses the CDK for 2D diagrams and descriptor calculation. This screenshot is provided by Kevin
+Lawson, Syngenta, UK, and reproduced with permission.},cap={},label={fig:tools},botcap,figure]
 {c}
@@
   \includegraphics[width=0.48\linewidth]{ambitOnline}
+  \includegraphics[width=0.48\linewidth]{DescriptorTest}
 }

### 2012-07-21 paper.tex
Addressed some of the review comments

@@
 generally require a numerical description of chemical structures. Such
-numerical forms range from a set of 3D coordinates (which coupled with
+numerical forms range from a set of 3D coordinates which coupled with
 appropriate atom types, is sufficient for methods such as quantum
@@
 is easy to understand \emph{why} two implementations of the same
-descriptor can lead to different results - the primary reasons being
-that the chemistry model that underlies the framework or toolkit used
-to implement the descriptor is different. For example, a descriptor
-that calculates the number of aromatic atoms may be implemented using
-two toolkits with differing aromaticity models, and hence it is
-possible that the values generated by the two implementations will
-differ. Other sources of differences include parameters that may be
-involved in the descriptor calculation and reference data values (such
-as atomic radii, electronegativity values) that are employed during
-descriptor calculation. While most implementations will employ the
-same data sources for standard concepts (e.g., atomic weights), slight
+descriptor can lead to different results. The primary reasons being
+differences in the chemistry model of the framework or toolkit used to
+implement the descriptor. For example, a descriptor that calculates
+the number of aromatic atoms may be implemented using two toolkits
+with differing aromaticity models, and hence it is possible that the
+values generated by the two implementations will differ. Other sources
+of differences include parameters that may be involved in the
+descriptor calculation and reference data values (such as atomic
+radii, electronegativity values) that are employed during descriptor
+calculation. While most implementations will employ the same data
+sources for standard concepts (e.g., atomic weights), slight
 differences in these types of input data can lead to differences in
@@
 \label{sec:categ-descr}
+
+\subsection{Numerical descriptors}
+\label{sec:numer-descr}
 
@@
 real valued fingerprint. As a result, the calculation procedure is
-rapid and once calculated shape similarities can be evaluated rapidly
+rapid and once calculated shape similarities can be evaluated quickly
 via a simple distance calculation. It should also be noted that shape
@@
 Puzyn et al question the need for such a high level of theory
-\cite{Puzyn:2008fk}).
+\cite{Puzyn:2008fk}). Sullivan et al \cite{Sullivan:2000fk} employed
+frontier orbital theory and DFT to generate quantum mechanical analogs
+of Hammett substituent constants \cite{Hammett:1937aa} and showed that
+they agreed well with empirically derived constants.
 
@@
 
-Every descriptor implements an interface allowing them to
-return values in a uniform manner, viz., via a \texttt{DescriptorValue}
-object (see Figure~\ref{fig:cdkapi}). This class can represents the result
-of a descriptor calculation and includes both the descriptor values
-(via a \texttt{IDescriptorResult} object) as well as meta-data on the
+Every descriptor class implements an interface allowing them to return
+values in a uniform manner, viz., via a \texttt{DescriptorValue}
+object. This class can represents the result of a descriptor
+calculation and includes both the descriptor values (via a
+\texttt{IDescriptorResult} object) as well as meta-data on the
 descriptor itself via a \texttt{DescriptorSpecification} object (see
@@
 implemented using SMARTS matching to identify the various paths and
-chains described by Kier \textit{et al.}, rather than explicitly walking the
-molecular graph.
+chains described by Kier \textit{et al.}, rather than explicitly
+walking the molecular graph.
 
@@
 manner. This makes automated comparisons and merging of descriptors
-from different sources problematic. To address the CDK has developed
+from different sources problematic. To address this, the CDK has developed
 the concept of a ``descriptor specification'' which is associated with
@@
 Molconn-Z & Commercial & No &  Yes & topological &  \NN
-OpenBabel & GPL & Yes & & \raggedright{} & \raggedright{} \NN
+OpenBabel & GPL & Yes & & \raggedright{constitutional} &
+\raggedright{MACCS, hashed} \NN
 PowerMV \cite{Liu:2005ab} & GPL &  & Yes & \raggedright{constitutional, atom pairs,
@@
 
-\ctable[caption={UML diagram of the CDK descriptor API in version 1.4.},cap={},label={fig:cdkapi},botcap,figure]
-{c}
-{}
-{
-  \includegraphics[width=\linewidth]{iDescriptor.pdf}
-}
+% \ctable[caption={UML diagram of the CDK descriptor API in version 1.4.},cap={},label={fig:cdkapi},botcap,figure]
+% {c}
+% {}
+% {
+%   \includegraphics[width=\linewidth]{iDescriptor.pdf}
+% }
 

### 2012-07-22 paper.tex
Made some more edits to address review

@@
 property over the molecular surface such as the Charged Partial
-Surface Area descriptors \cite{Stanton:1990aa} that characterize the
-partial charge distribution over the molecular surface. Note that
-surface descriptor calculations based on analytical
+Surface Area descriptors (CPSA) \cite{Stanton:1990aa} that
+characterize the partial charge distribution over the molecular
+surface. Note that surface descriptor calculations based on analytical
 \cite{Connolly:1983aa} or tessellation algorithms
@@
 such as the Topological Polar Surface Area (TPSA) \cite{Ertl:2000aa}
-and VSA descriptors \cite{Labute:2008aa} that are based on
-precalculated surface area values derived from a list of functional
-groups.
+and MOE van der Waals Surface Area (VSA) descriptors
+\cite{Labute:2008aa} that are based on precalculated surface area
+values derived from a list of functional groups.
 
@@
 categories are based on input requirements. This gives rise to
-situations where the molecular volume property can be
-described by geometrical descriptors, e.g. using an analytical
-approaches ~\cite{Eisenhaber:1995qf}, and with approximate,
-constitutional descriptors, such as the very fast algorithm use by the
-VABC approach~\cite{Zhao:2003fk} which is based on group
-contributions.  The latter descriptor has been implemented in the CDK
-and correlates well with other approximations to the molecular volume
-(Figure \ref{fig:molvol}).
+situations where the molecular volume property can be described by
+geometrical descriptors, e.g. using an analytical approaches
+~\cite{Eisenhaber:1995qf}, and with approximate, constitutional
+descriptors, such as the very fast algorithm use by the van der Waals
+Volume as a Sum of Atomic and Bond Contributions (VABC)
+approach~\cite{Zhao:2003fk} which is based on group contributions.
+The latter descriptor has been implemented in the Chemistry
+Development Kit (CDK) and correlates well with other approximations to
+the molecular volume (Figure \ref{fig:molvol}).
 
 Another important class of geometrical descriptors are those that
-describe molecular shape. This property is a key factor in ligand-receptor interactions and is
-an important approach to virtual screening. A variety of descriptors
-have been devised to characterize molecular shape. The use of
-overlapping atom centered Gaussians to represent molecular shape
-\cite{Grant:1995aa} represents a significant computational advantage
-over the traditional analytical approach of intersecting hard spheres,
-supporting extremely rapid shape comparisons \cite{Grant:1999vn}. An
-alternative approach described by Ritchie et al \cite{Ritchie:1999kx}
-employed a spherical harmonic based representation of surfaces, that
-in contrast to the Gaussian approach allows one to generate
-characterizations of the surface and hence the shape at varying
-resolutions. Another approach to multi-resolution shape
-representations is the use of $\alpha$ shapes
+describe molecular shape. This property is a key factor in
+ligand-receptor interactions and is an important approach to virtual
+screening. A variety of descriptors have been devised to characterize
+molecular shape. The use of overlapping atom centered Gaussians to
+represent molecular shape \cite{Grant:1995aa} represents a significant
+computational advantage over the traditional analytical approach of
+intersecting hard spheres, supporting extremely rapid shape
+comparisons \cite{Grant:1999vn}. An alternative approach described by
+Ritchie et al \cite{Ritchie:1999kx} employed a spherical harmonic
+based representation of surfaces, that in contrast to the Gaussian
+approach allows one to generate characterizations of the surface and
+hence the shape at varying resolutions. Another approach to
+multi-resolution shape representations is the use of $\alpha$ shapes
 \cite{Edelsbrunner:1994aa}, which represent a parameterization of the
@@
 
-The last category consists of QM descriptors. These
-descriptors can also be applied to problems such as QSAR models
-(though of course the time complexity of QM derived descriptors do not
-make them amenable for large scale virtual screening
-applications). These descriptors can be evaluated at various levels of
-theory. For example, partial charges, HOMO and LUMO energies are
-commonly evaluated using semi-empirical methods, while Eroglu and
-T\"{u}rkmen \cite{Eroglu:2007uq} described DFT-derived descriptors
-such as electronegativity, hardness and ionization potential (though
-Puzyn et al question the need for such a high level of theory
-\cite{Puzyn:2008fk}). Sullivan et al \cite{Sullivan:2000fk} employed
-frontier orbital theory and DFT to generate quantum mechanical analogs
-of Hammett substituent constants \cite{Hammett:1937aa} and showed that
-they agreed well with empirically derived constants.
+The last category consists of QM descriptors. These descriptors can
+also be applied to problems such as Quantitative Structure-Activity
+Relationship (QSAR) models (though of course the
+time complexity of QM derived descriptors do not make them amenable
+for large scale virtual screening applications). These descriptors can
+be evaluated at various levels of theory. For example, partial
+charges, Highest Occupied Molecular Orbital (HOMO) and Lowest Occupied
+MOlecular Orbital (LUMO) energies are commonly evaluated using
+semi-empirical methods, while Eroglu and T\"{u}rkmen
+\cite{Eroglu:2007uq} described density functional theory (DFT) derived
+descriptors such as electronegativity, hardness and ionization
+potential (though Puzyn et al question the need for such a high level
+of theory \cite{Puzyn:2008fk}). Sullivan et al \cite{Sullivan:2000fk}
+employed frontier orbital theory and DFT to generate quantum
+mechanical analogs of Hammett substituent constants
+\cite{Hammett:1937aa} and showed that they agreed well with
+empirically derived constants.
 
@@
 descriptors generally tend to be based on the individual amino
-acids. Examples include the VHSE family of descriptors described by
-Mei et al \cite{Mei:2005kx}, a set of side chain descriptors from
-Collantes and Dunn \cite{Collantes:1995vn}, field based descriptors
+acids. Examples include the Vector of Hydrophobic, Steric, and
+Electronic (VHSE) family of descriptors described by Mei et al
+\cite{Mei:2005kx}, a set of side chain descriptors from Collantes and
+Dunn \cite{Collantes:1995vn}, field based descriptors
 \cite{Norinder:1991ys} and the AAindex database
@@
 Sedykh et al \cite{Sedykh:2011fk} who hypothesized that dose response
-data points from high throughput dose-response assays could be employed
-as biological descriptors. They observed that when such descriptors
-were combined with traditional chemical structure descriptors, the
-predictive performance of models developed using the combined set was
-better than those developed using conventional approaches. A similar
-approach can be taken using the PASS methodology
+data points from high throughput dose-response assays could be
+employed as biological descriptors. They observed that when such
+descriptors were combined with traditional chemical structure
+descriptors, the predictive performance of models developed using the
+combined set was better than those developed using conventional
+approaches. A similar approach can be taken using the Prediction of
+Biological Activity Spectra for Chemical Substances (PASS) methodology
 \cite{Poroikov:2007aa}, though this is an indirect method - one must
@@
 (\url{http://rguha.net/code/java/cdkdesc.html}) is a Java Swing
-application that exposes the Chemistry Development Kit (CDK) descriptor and fingerprint API's. It
+application that exposes the CDK descriptor and fingerprint API's. It
 allows the users to specify a SMILES or SD file and evaluate a
@@
 applications that allow easy descriptor calculations. Examples include
-Bioclipse \cite{Spjuth:2007aa} (see Figure~\ref{fig:tools}), MOE (Chemical Computing Group) and
-Maestro (Schr\"{o}dinger, Inc.) as standalone tools, but also
-as a plugin to Microsoft Excel, called LICSS~\cite{Lawson2011}.
-Workflow tools are also a useful
-class of applications for descriptor generation and allow one to
-easily generate descriptors from multiple sources~\cite{Truszkowski2011}
-(see Figure~\ref{fig:tools}). Given the focus of
-this article on Open Source implementations, we have noted the
-availability in Table \ref{tab:impl}. While a number of
-implementations are not strictly Open Source according to OSI
-definitions, the fact that they provide free academic licenses does
-allow them to be used somewhat freely.
+Bioclipse \cite{Spjuth:2007aa} (see Figure~\ref{fig:tools}), MOE
+(Chemical Computing Group) and Maestro (Schr\"{o}dinger, Inc.) as
+standalone tools, but also as a plugin to Microsoft Excel, called
+LICSS~\cite{Lawson2011}.  Workflow tools are also a useful class of
+applications for descriptor generation and allow one to easily
+generate descriptors from multiple sources~\cite{Truszkowski2011} (see
+Figure~\ref{fig:tools}). Given the focus of this article on Open
+Source implementations, we have noted the availability in Table
+\ref{tab:impl}. While a number of implementations are not strictly
+Open Source according to OSI definitions, the fact that they provide
+free academic licenses does allow them to be used somewhat freely.
 
@@
 refer to descriptors via a standardized naming scheme. Currently the
-QSAR modeling community lacks such a scheme, though the CDK, a Java library for cheminformatics
-\cite{Steinbeck:2006aa} does implement such a scheme. The ``descriptor
-specification'' approach defines a set of classes that include
-information of the implementation title, identifier and vendor. The
-identifier may or may not include a version number. The actual details
-of the a descriptor are listed in a dictionary, along with an optional
-namespace. The dictionary entry is referenced by the descriptor
-specification and this design thus allows one to seamlessly work with
-multiple implementations of the same descriptor from different vendors
-and keep track of which version is employed throughout a calculation
-or workflow. In addition, the CDK provides a hierarchical annotation
-of descriptors, grouping them into different classes. Since this
-annotation is based on RDF \cite{Taylor:2006ab}, it allows one to go
+QSAR modeling community lacks such a scheme, though the CDK, a Java
+library for cheminformatics \cite{Steinbeck:2006aa} does implement
+such a scheme. The ``descriptor specification'' approach defines a set
+of classes that include information of the implementation title,
+identifier and vendor. The identifier may or may not include a version
+number. The actual details of the a descriptor are listed in a
+dictionary, along with an optional namespace. The dictionary entry is
+referenced by the descriptor specification and this design thus allows
+one to seamlessly work with multiple implementations of the same
+descriptor from different vendors and keep track of which version is
+employed throughout a calculation or workflow. In addition, the CDK
+provides a hierarchical annotation of descriptors, grouping them into
+different classes. Since this annotation is based on the Resource
+Description Framework (RDF) \cite{Taylor:2006ab}, it allows one to go
 beyond simple groupings and perform reasoning on descriptors. A simple
@@
 cheminformatics~\cite{Hastings2011}, which is used by the CHESS
-framework which also uses the CDK for descriptor calculations~\cite{Chepelev2011b}.
+framework which also uses the CDK for descriptor
+calculations~\cite{Chepelev2011b}.
 
@@
 underlying chemistry models and reference data used by the
-implementations. As shown by Gupta et al \cite{Gupta:2010uq},
-the performance of predictive models developed using commercial or
-Open Source descriptor implementations is very similar. While
-different tools differ in the specific number and types of descriptors
-that they calculate, the fact many descriptors are correlated with
-others suggests that it does not matter too much which set of
-descriptors are used in an application. Of course, this cannot be a
-general rule - specific applications may require a specific set of
-descriptors. This is not to say that all descriptor tools exhibit
-similar performance. For example, it is clear from Figure \ref{fig:logp}
-that the ACD Labs implementation of log P performs significantly
-better than other implementations. Though this is not a completely
-rigorous comparison (the methods underlying the ACD Labs, CDK and
-ChemAxon implementations are different), it does highlight that
-certain implementations of a descriptor can fare better than others -
-especially those cases where the descriptor is based on a predictive
-model.
+implementations. As shown by Gupta et al \cite{Gupta:2010uq}, the
+performance of predictive models developed using commercial or Open
+Source descriptor implementations is very similar. While different
+tools differ in the specific number and types of descriptors that they
+calculate, the fact many descriptors are correlated with others
+suggests that it does not matter too much which set of descriptors are
+used in an application. Of course, this cannot be a general rule -
+specific applications may require a specific set of descriptors. For
+example, a predictive model for boiling points may not peform very
+well if branching and intermolecular forces are not taken into account
+by the descriptors in the model. This is not to say that all
+descriptor tools exhibit similar performance. For example, it is clear
+from Figure \ref{fig:logp} that the ACD Labs implementation of log P
+performs significantly better than other implementations. Though this
+is not a completely rigorous comparison (the methods underlying the
+ACD Labs, CDK and ChemAxon implementations are different), it does
+highlight that certain implementations of a descriptor can fare better
+than others - especially those cases where the descriptor is based on
+a predictive model.
 

### 2012-07-24 paper.tex
Added more to rebuttal

@@
 same information, allowing one to be substituted for another. The
-choice of descriptors is a well-known problem and given a large
-collection of them, approaches to identify a suitable subset have been
-discussed extensively in the
+selection of relevant descriptors is a well-known problem and given a
+large collection of them, approaches to identify a suitable subset
+have been discussed extensively in the
 literature~\cite{Miller:2002aa,Kohavi:1997gf}.

### 2012-07-24 paper.tex
Added a nice intro figure to the paper
made sure to refer to the figure in the text;

@@
 descriptions derived from 2D or 3D representations which can be useful
-in statistical approaches. It is now possible to evaluate thousands of
-numerical descriptors of chemical structure. As will be discussed
+in statistical approaches.  It is now possible to evaluate thousands
+of numerical descriptors of chemical structure. As will be discussed
 later, many of these descriptors are closely related or capture the
@@
 have been discussed extensively in the
-literature~\cite{Miller:2002aa,Kohavi:1997gf}.
+literature~\cite{Miller:2002aa,Kohavi:1997gf}.  Figure \ref{fig:repr}
+is a summary depiction of the major types of descriptors and the form
+of molecular structure information that is required to compute
+them. The depiction is very general and focuses on small molecule
+descriptors. As will be described in the following sections, molecular
+descriptors can be calculated for many chemical entities, not just
+small organic molecules.
 
@@
 
+\ctable[caption={A graphical summary of descriptor types and the type
+  of input information required. As one goes from top to bottom the
+calculations become more intensive, but the results capture aspects of
+molecular structure more realistically.}, cap={}, botcap,figure, label={fig:repr}]
+{c}
+{}
+{
+\includegraphics[width=0.9\linewidth]{oss-desc-repr}
+}
+
 \ctable[caption={A comparison of the VABC descriptor for molecular

### 2012-07-24 paper.tex
some more response to reviews

@@
 corresponding to small molecule structures (even though they may be
-computationally intractable on much larger graphs). While this class
-of descriptors have been used extensively in predictive modeling
-applications
-\cite{Garcia-Domenech:2008aa,Randic:2001ad,Besalu:2001aa}, a key
-drawback is their lack of interpretability. The abstract nature of
-many graph invariants and information theoretic descriptors, their use
-in predictive models makes it difficult to explain the model in simple
-physicochemical terms, though various attempts have been made at
-interpretations \cite{Todeschini:1975dq,Stanton:2003aa}.
+computationally intractable on much larger graphs). An example of a
+topological descriptor is the Wiener index
+\cite{Nikolic:2001aa,Wiener:1947aa}. It is simply the sum of the edge
+counts in the shortest paths between all pairs of non-hydrogen
+atoms. From a physical standpoint, the descriptor characterizes the
+branching in a molecular structure. While this class of descriptors
+have been used extensively in predictive modeling applications
+\cite{Garcia-Domenech:2008aa,Randic:2001ad,Besalu:2001aa} (for
+example, the Wiener index described above has been useful in models of
+boiling point data), a key drawback is their general lack of
+interpretability. The abstract nature of many graph invariants and
+information theoretic descriptors, their use in predictive models
+makes it difficult to explain the model in simple physicochemical
+terms, though various attempts have been made at interpretations
+\cite{Todeschini:1975dq,Stanton:2003aa}.
 

### 2012-07-24 paper.tex
updated plots

@@
   volume (as implemented by the CDK) and other approximate
-  implementations of molecular volume (as implemented by MOE 2008.10).},cap={},label={fig:molvol},botcap,figure]
+  implementations of molecular volume (as implemented by MOE
+  2008.10). Both plots are overlaid with a loess fit.},cap={},label={fig:molvol},botcap,figure]
 {c}

### 2012-07-24 code/code.R

@@ -1,5 +1,6 @@
 library(lattice)
 library(reshape)
+library(ggplot2)
 
 acd <- read.table('all-clean.acd', sep='\t', header=TRUE, as.is=TRUE)
 cdk <- read.table('all-clean.cdk', sep='\t', header=TRUE, as.is=TRUE)
@@ -20,11 +21,19 @@ dev.copy(png, '../tpsa-cdk-acd.png', 400,400);dev.off()
 ## volumes
 moe <- read.csv('vol.moe', header=TRUE, as.is=TRUE)
 cdk <- read.table('vol.cdk', header=TRUE, as.is=TRUE, sep=' ')
-names(moe) <- c('sid', 'vdw_vol', 'vol')
+names(moe) <- c('sid', 'van der Waals Volume (2D)', 'van der Waals Volume (grid based)')
 names(cdk) <- c('sid', 'VABC')
 tmp <- merge(moe, cdk, by='sid')
 tmp <- tmp[ complete.cases(tmp), ]
-tmp <- melt(tmp, measure.vars=c('vdw_vol', 'vol'), id.vars='VABC')
+tmp <- melt(tmp,, id.vars='VABC', measure.vars=c('van der Waals Volume (2D)', 'van der Waals Volume (grid based)'))
+
+ggplot(tmp, aes(x=VABC,y=value))+
+  geom_point(size=1,alpha=0.25)+
+  geom_smooth()+
+  facet_wrap(~variable)+
+  scale_x_continuous(limits=c(0,1500))+
+  xlab('CDK VABC')+ylab("MOE")
+
 xyplot(value ~ VABC | variable, tmp,
        type=c('p', 'g'), col='black', cex=0.25,
        xlab=list(label='CDK VABC', cex=1.2),

### 2012-07-24 code/code.R
some more fig updates

@@ -4,8 +4,14 @@ library(ggplot2)
 
 acd <- read.table('all-clean.acd', sep='\t', header=TRUE, as.is=TRUE)
 cdk <- read.table('all-clean.cdk', sep='\t', header=TRUE, as.is=TRUE)
-
 tmp <- data.frame(ACD=acd$ACD_RuleOf5_PSA, CDK=cdk$TopoPSA)
+
+ggplot(tmp, aes(x=CDK,y=ACD))+
+  geom_point(size=1.5,alpha=0.25)+
+  geom_smooth()+
+#  scale_x_continuous(limits=c(0,1500))+
+  xlab('CDK TPSA')+ylab("ACD TPSA")
+
 xyplot(ACD ~ CDK, tmp, type=c('p', 'g'), col='black', cex=0.25,
        xlab=list(label='CDK TPSA', cex=1.2),
        ylab=list(label='ACD TPSA', cex=1.2),

### 2012-07-29 paper.tex
Added missing units
Fixed two typos

@@
 specific applications may require a specific set of descriptors. For
-example, a predictive model for boiling points may not peform very
+example, a predictive model for boiling points may not perform very
 well if branching and intermolecular forces are not taken into account
@@
 in Ref.~\citen{Todeschini:2002aa}} &  \NN
-JChem & Commercial & Yes & Yes & \raggedright{Consititutional,
+JChem & Commercial & Yes & Yes & \raggedright{Constitutional,
   geometric, topological, electronic, hybrid} & \raggedright{Hashed,
@@
   implementations of molecular volume (as implemented by MOE
-  2008.10). Both plots are overlaid with a loess fit.},cap={},label={fig:molvol},botcap,figure]
+  2008.10). All values are in \AA$^{3}$ and both plots are overlaid with a loess fit.},cap={},label={fig:molvol},botcap,figure]
 {c}

### 2012-07-29 paper.tex
final set of edits

@@
 
-Given that we can generate thousands of descriptors,  it is critical to ask, what
-makes a descriptor useful? Fundamentally, a descriptor must correlate
-structural features with some physicochemical property and show
-minimal correlation with other descriptors. In addition, a generally
-useful descriptor will be applicable to a broad class of
+Given that we can generate thousands of descriptors, it is critical to
+ask, what makes a descriptor useful? Fundamentally, a descriptor must
+correlate structural features with some physicochemical property and
+show minimal correlation with other descriptors. In addition, a
+generally useful descriptor will be applicable to a broad class of
 molecules. Furthermore, a descriptor should show minimal
@@
 differences are small. An example is the case of stereochemistry. Most
-constitutional and topological descriptors that are not designed to take this into account will
-generate the same numerical values for all the stereoisomers of a
-given chiral molecule. We would consider such a descriptor to be
-degenerate. In addition to degeneracy, a descriptor should be
-continuous, in the sense, that small structural changes should lead to
-small changes in the value of the descriptor.
+constitutional and topological descriptors that are not designed to
+take this into account will generate the same numerical values for all
+the stereoisomers of a given chiral molecule. We would consider such a
+descriptor to be degenerate. In addition to degeneracy, a descriptor
+should be continuous, in the sense, that small structural changes
+should lead to small changes in the value of the descriptor.
 
 The requirements listed here ensures that subsequent analyses
-involving descriptors are robust. But in addition to numerical
-robustness, it is important for many scenarios that the descriptor
+involving descriptors are robust. But there are other aspects of a
+descriptor that can make it more or less useful. For example,a
+descriptor that can be rapidly calculated and is not dependent on
+experimental properties can be considered more useful than one that is
+computationally intensive and / or dependent on experimental
+results. In addition, it is important for many scenarios that the descriptor
 have some form of physical interpretability.  For example, a model
@@
 descriptors.
+
+This discussion has focused on aspects of descriptors that make them
+useful in computational models. It is also relevant to note that the
+utility of a descriptor can be dependent on the problem being
+considered. More specifically, certain end points may require specific
+molecular features to be taken into account. Descriptors that do not,
+will not be useful in developing models for that endpoint. For
+example, an accurate boiling point model will require descriptors that
+capture information regarding molecular branching, hydrogen bonding
+potential and so on. Similarly, a model to predict activity against a
+cytoplasmic or nuclear target, will likely require that molecular
+characteristics enabling passage across the cell membrane be taken
+into account. In such a case, a descriptor characterizing the
+lipophilicity of the molecule will be useful. These points also apply
+to other types of computational approaches, such as clustering. In the
+end, the utility of a descriptor is determined by whether it encapsulates
+sufficient and relevant information. In general a single descriptor
+will not satisfy these requirements and thus one usually works with
+sets of descriptors.
 
@@
   volume (as implemented by the CDK) and other approximate
-  implementations of molecular volume (as implemented by MOE
-  2008.10). All values are in \AA$^{3}$ and both plots are overlaid with a loess fit.},cap={},label={fig:molvol},botcap,figure]
+  implementations of molecular volume (a 2D approach and a grid based
+  approach), as implemented by MOE
+  2008.10. All values are in \AA$^{3}$ and both plots are overlaid with a loess fit.},cap={},label={fig:molvol},botcap,figure]
 {c}
